---
title: "HW2"
author: "Mengyan Jing and Abhishek Banerjee"
date: "2026-02-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1


```{r}
integrand <- function(t, s) {
  # Use log-scale computation to avoid overflow/underflow
  exp((s - 1) * log(t) - t)
}

lower_inc_gamma <- function(s, x) {
  if (x <= 0) return(0)
  
  # For large x, compute as gamma(s) - upper
  if (x > 700) {
    return(gamma(s) - upper_inc_gamma(s, x))
  }
  
  # Split: integrate from a small epsilon to x
  # Near 0, use substitution u = t^s to remove singularity
  # Original: integral of t^(s-1)*exp(-t) dt from 0 to x
  # Sub u = t^s => t = u^(1/s), dt = (1/s)*u^(1/s - 1) du
  # Integral becomes (1/s) * integral of exp(-u^(1/s)) du from 0 to x^s
  
  if (s < 1) {
    # Use substitution
    transformed_integrand <- function(u, s) {
      (1/s) * exp(-u^(1/s))
    }
    result <- integrate(transformed_integrand, lower = 0, upper = x^s, s = s,
                        rel.tol = 1e-10, subdivisions = 1000L)
  } else {
    result <- integrate(integrand, lower = 0, upper = x, s = s,
                        rel.tol = 1e-10, subdivisions = 1000L)
  }
  return(result$value)
}

upper_inc_gamma <- function(s, x) {
  if (x <= 0) return(gamma(s))
  result <- integrate(integrand, lower = x, upper = Inf, s = s,
                      rel.tol = 1e-10, subdivisions = 1000L)
  return(result$value)
}

lower_inc_gamma_vec <- function(s_vec, x_vec) {
  mapply(lower_inc_gamma, s_vec, x_vec)
}

upper_inc_gamma_vec <- function(s_vec, x_vec) {
  mapply(upper_inc_gamma, s_vec, x_vec)
}
```

```{r}
s_vals <- seq(0.0001, 10, length.out = 100)
x_vals <- seq(0.0001, 10000, length.out = 100)
param_set <- expand.grid(s = s_vals, x = x_vals)

# Time lower
ptm <- proc.time()
lower_results <- lower_inc_gamma_vec(param_set$s, param_set$x)
proc.time() - ptm

# Time upper
ptm <- proc.time()
upper_results <- upper_inc_gamma_vec(param_set$s, param_set$x)
proc.time() - ptm

```


```{r}
# Baseline from R's built-in functions
baseline_lower <- pgamma(param_set$x, shape = param_set$s, scale = 1, 
                          lower.tail = TRUE) * gamma(param_set$s)
baseline_upper <- pgamma(param_set$x, shape = param_set$s, scale = 1, 
                          lower.tail = FALSE) * gamma(param_set$s)

# MSE
mse_lower <- mean((lower_results - baseline_lower)^2)
mse_upper <- mean((upper_results - baseline_upper)^2)

cat("MSE lower:", mse_lower, "\n")
cat("MSE upper:", mse_upper, "\n")

```


```{r}
library(Rcpp)
```

```{r}
Sys.which("make")
```

## Task 1

```{Rcpp}
#include <Rcpp.h>
#include <cmath>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector integrand_cpp(NumericVector t, double s) {
  int n = t.size();
  NumericVector out(n);
  for(int i=0; i<n; i++){
    out[i] = std::exp((s - 1.0) * std::log(t[i]) - t[i]);
  }
  return out;
}

// [[Rcpp::export]]
NumericVector transformed_integrand_cpp(NumericVector u, double s) {
  int n = u.size();
  NumericVector out(n);
  for(int i=0; i<n; i++){
    out[i] = (1.0 / s) * std::exp(-std::pow(u[i], 1.0 / s));
  }
  return out;
}
```



```{r}
lower_inc_gamma_cpp <- function(s, x) {
  if (x <= 0) return(0)
  
  if (x > 700) {
    return(gamma(s) - upper_inc_gamma_cpp(s, x))
  }
  
  if (s < 1) {
    result <- integrate(function(u) transformed_integrand_cpp(u, s),
                        lower = 0, upper = x^s,
                        rel.tol = 1e-10, subdivisions = 1000L)
  } else {
    result <- integrate(function(t) integrand_cpp(t, s),
                        lower = 0, upper = x,
                        rel.tol = 1e-10, subdivisions = 1000L)
  }
  return(result$value)
}

upper_inc_gamma_cpp <- function(s, x) {
  if (x <= 0) return(gamma(s))
  result <- integrate(function(t) integrand_cpp(t, s),
                      lower = x, upper = Inf,
                      rel.tol = 1e-10, subdivisions = 1000L)
  return(result$value)
}

lower_inc_gamma_cpp_vec <- function(s_vec, x_vec) {
  mapply(lower_inc_gamma_cpp, s_vec, x_vec)
}

upper_inc_gamma_cpp_vec <- function(s_vec, x_vec) {
  mapply(upper_inc_gamma_cpp, s_vec, x_vec)
}
```


```{r}
# Timing
ptm <- proc.time()
lower_results_cpp <- lower_inc_gamma_cpp_vec(param_set$s, param_set$x)
proc.time() - ptm

ptm <- proc.time()
upper_results_cpp <- upper_inc_gamma_cpp_vec(param_set$s, param_set$x)
proc.time() - ptm

# MSE
mse_lower_cpp <- mean((lower_results_cpp - baseline_lower)^2)
mse_upper_cpp <- mean((upper_results_cpp - baseline_upper)^2)
cat("MSE lower (C++ integrand):", mse_lower_cpp, "\n")
cat("MSE upper (C++ integrand):", mse_upper_cpp, "\n")
```


## 3)

```{r}
# install.packages("statmod")
library(statmod)

gw <- gauss.quad(n = 30, kind = "legendre")
nodes <- gw$nodes
weights <- gw$weights

nodes
weights
```


```{Rcpp}
#include <Rcpp.h>
#include <cmath>
#include <vector>
using namespace Rcpp;

// 30-point Gauss-Legendre nodes and weights
void gauss_legendre_30(std::vector<double>& nodes, std::vector<double>& weights) {
  const double x[] = {
    0.0514718426, 0.1538699136, 0.2546369261, 0.3527047255, 0.4470337695,
    0.5366241481, 0.6205261830, 0.6978504948, 0.7677774321, 0.8295657624,
    0.8825605358, 0.9262000474, 0.9600218650, 0.9836681233, 0.9968934841
  };
  const double w[] = {
    0.1028526529, 0.1017623897, 0.0995934206, 0.0963687372, 0.0921225222,
    0.0868997872, 0.0807558952, 0.0737559747, 0.0659742299, 0.0574931562,
    0.0484026728, 0.0387991926, 0.0287847079, 0.0184664683, 0.0079681925
  };
  nodes.resize(30);
  weights.resize(30);
  for (int i = 0; i < 15; i++) {
    nodes[i]        = -x[14 - i];
    weights[i]      =  w[14 - i];
    nodes[15 + i]   =  x[i];
    weights[15 + i] =  w[i];
  }
}

static std::vector<double> gl_nodes, gl_wts;
static bool gl_init = false;

void ensure_gl() {
  if (!gl_init) {
    gauss_legendre_30(gl_nodes, gl_wts);
    gl_init = true;
  }
}

// Generic GL integration over [a,b] with n_panels
typedef double (*integrand_fn)(double, double, double);

double gl_integrate2(double a, double b, int n_panels,
                     integrand_fn f, double p1, double p2) {
  ensure_gl();
  double h = (b - a) / n_panels;
  double total = 0.0;
  for (int p = 0; p < n_panels; p++) {
    double mid = a + h * (p + 0.5);
    double half_h = h * 0.5;
    for (int i = 0; i < 30; i++) {
      double t = mid + half_h * gl_nodes[i];
      total += half_h * gl_wts[i] * f(t, p1, p2);
    }
  }
  return total;
}

// Direct integrand: t^(s-1)*exp(-t)
double direct_ig(double t, double s, double unused) {
  if (t <= 0.0) return 0.0;
  return std::exp((s - 1.0) * std::log(t) - t);
}

// Transformed for s<1: (1/s)*exp(-u^(1/s))
double transformed_ig(double u, double s, double unused) {
  if (u <= 0.0) return 0.0;
  return (1.0 / s) * std::exp(-std::pow(u, 1.0 / s));
}

// NEW: stable upper integrand on v ∈ [0, Vmax], with t = x + v
double upper_v_ig(double v, double s, double x) {
  double t = x + v;
  if (t <= 0.0) return 0.0;
  return std::exp((s - 1.0) * std::log(t) - t);
}

// Panels heuristic
int choose_panels(double range) {
  if (range < 1.0) return 20;
  if (range < 10.0) return 40;
  if (range < 100.0) return 80;
  if (range < 1000.0) return 160;
  return 300;
}

//
double lower_scalar(double s, double x) {
  if (x <= 0.0) return 0.0;

  if (s < 1.0) {
    double upper_lim = std::pow(x, s);
    int panels = choose_panels(upper_lim);
    return gl_integrate2(1e-15, upper_lim, panels, transformed_ig, s, 0.0);
  } else {
    int panels = choose_panels(x);
    return gl_integrate2(1e-15, x, panels, direct_ig, s, 0.0);
  }
}

double upper_scalar(double s, double x) {
  if (x <= 0.0) return R::gammafn(s);

  // very large x => ~0
  double cutoff = s + 50.0 * std::sqrt(s) + 200.0;
  if (x > cutoff) return 0.0;

  // Key fix: for x < s, use complement (avoids missing mass)
  if (x < s + 2.0 * std::sqrt(s)) {
    return R::gammafn(s) - lower_scalar(s, x);
  }

  // Otherwise compute upper directly, stable v-integral (no 1/u)
  double Vmax = 60.0 + 10.0 * std::sqrt(s);   // baseline
  if (x > 10.0) Vmax = 40.0 + 8.0 * std::sqrt(s);
  if (x > 50.0) Vmax = 25.0 + 5.0 * std::sqrt(s);   // safe truncation
  int panels = choose_panels(Vmax);
  return gl_integrate2(0.0, Vmax, panels, upper_v_ig, s, x);
}

// [[Rcpp::export]]
NumericVector lower_inc_gamma_full_cpp(NumericVector s_vec, NumericVector x_vec) {
  int n = s_vec.size();
  NumericVector out(n);

  for (int i = 0; i < n; i++) {
    double s = s_vec[i];
    double x = x_vec[i];

    if (x <= 0.0) { out[i] = 0.0; continue; }

    // For huge x, lower = gamma(s) - upper (upper tiny)
    if (x > 700.0) {
      out[i] = R::gammafn(s) - upper_scalar(s, x);
    } else {
      out[i] = lower_scalar(s, x);
    }
  }
  return out;
}

// [[Rcpp::export]]
NumericVector upper_inc_gamma_full_cpp(NumericVector s_vec, NumericVector x_vec) {
  int n = s_vec.size();
  NumericVector out(n);

  for (int i = 0; i < n; i++) {
    out[i] = upper_scalar((double)s_vec[i], (double)x_vec[i]);
  }
  return out;
}
```



```{r}
# Timing
ptm <- proc.time()
lower_results_full <- lower_inc_gamma_full_cpp(param_set$s, param_set$x)
proc.time() - ptm

ptm <- proc.time()
upper_results_full <- upper_inc_gamma_full_cpp(param_set$s, param_set$x)
proc.time() - ptm

# MSE
mse_lower_full <- mean((lower_results_full - baseline_lower)^2)
mse_upper_full <- mean((upper_results_full - baseline_upper)^2)
cat("MSE lower (full C++):", mse_lower_full, "\n")
cat("MSE upper (full C++):", mse_upper_full, "\n")
```

# Problem 3

We have in the question, an exponentially tilted positive stable distribution which is very useful and flexible in modeling extreme weather. The family of such distributions is parameterized by $\theta > 0$. We have its density function as:
$$
f(x;\theta)
=
\frac{\sqrt{\frac{1}{4\pi x^{3}}}\,
\exp\!\left(-\frac{1}{4x} - \theta x\right)}
{\exp(-\sqrt{\theta})},
\qquad x > 0.
$$
## 1)

We will use Rejection Sampling technique discueesed in class for this problem. If we look at the above density and if we ignore the constant terms, we get the kernal as $$x^{-3/2} \exp\!\left(-\frac{1}{4x} - \theta x\right)$$. This kernal, at least to us, looks strikingly similar to that of the kernel of an Inverse Gaussian distribution. From wikipedia, I got the density of Inverse Gaussian distribution as $$
f(x;\mu,\lambda)
=
\sqrt{\frac{\lambda}{2\pi x^{3}}}
\exp\!\left(
-\frac{\lambda (x-\mu)^2}{2\mu^{2}x}
\right),
\qquad x>0,$$ where $\mu >0$ and $\lambda > 0$ are parameters. Thus, we can use this distribution as a proposal distribution. We can use in-built R function to sample from this proposal distribution.  We use the following R code to implement the Rejection sampling with Inverse Gaussian being our proposal distribution. 

```{r}
library(statmod)   

# Target density
f_theta <- function(x, theta = 0.1){
  ifelse(x > 0,
         sqrt(1/(4*pi*x^3)) *
           exp(-1/(4*x) - theta*x + sqrt(theta)),
         0)
}

# Proposal: Inverse Gaussian
g_prop <- function(x, theta = 0.1){
  mu  <- 1/(2*sqrt(theta))
  lam <- 1
  dinvgauss(x, mean = mu, shape = lam)
}

# Rejection sampler 
tilted_sampler <- function(theta = 0.1){

  mu  <- 1/(2*sqrt(theta))
  lam <- 1

  counter <- 1

  x <- rinvgauss(1, mean = mu, shape = lam)
  ratio <- f_theta(x, theta) / g_prop(x, theta)

  while(runif(1) > ratio){

    x <- rinvgauss(1, mean = mu, shape = lam)
    counter <- counter + 1

    ratio <- f_theta(x, theta) / g_prop(x, theta)
  }

  return(c(x, counter))
}
```

Now as asked in the question, we sample $N=100000$ independent samples for $\theta= 0.1$


```{r}
N <- 100000
res <- sapply(1:N, function(i) tilted_sampler(0.1))

samples     <- res[1,]
num_sampled <- sum(res[2,])
```


## 2) 

Now we replicate Figure 3 in the question; that is, plot the sample density using histogram and compare it to the underlying true density. We use the ocde given in the question:

```{r}
hist(samples, breaks = 1000, freq = FALSE,
     border = "lightblue")

curve(f_theta(x, 0.1),
      from = 1e-5, to = 60,
      n = 1000, add = TRUE,
      col = "red", lty = 3)
```

We see in the above plot, there is a perfect alignment between the sample density and the analytic density. This showed that our choice of proposal distribution was correct and the rejection sampling technique henceforth did a good job!



# Problem 4 

In this question we consider a 4 × 4 grid of Bernoulli random variables denoted by $X = [Xij]$. We also consider $N(X)$ as the total number of successes in X (the sum of X) and $D(X)$ as the total number of neighbors (horizontal or vertical) which differ. We are also told that if we write the elements of X out in a vector V , then there is a 24×16 matrix M such that $D(X)$ is sum of the absolute values of MV . The 24 arises because there are 24 neighbor-pairs to check (diagonal elements are not neighbors.) We have probability distribution of X being proportional to $$
\pi(X) \propto 
p^{N(X)} (1 - p)^{16 - N(X)} 
\exp\!\left\{ -\lambda D(X) \right\}.$$
Therefore, if $\lambda = 0$, we have i.i.d.\ Bernoulli entries. We use Markov chain Monte Carlo to approximate the probability that the diagonal of this matrix is all 1’s. We assume $p = 0.8$. We approximate the probabilities for each of the following cases: $\lambda = 0.5, \: \lambda = 1.$ We have $Q1$ where every transition is equally likely with probability $1/2^{16}$. That is, the next candidate state is simply 16 i.i.d. Bernoulli variables with prob- ability 0.5 of success each. We use $Q2$ such that we choose one of the 16 entries in the table with probability 1/16 and then set that entry to 0 or 1 with probability 0.5 each. We use the following R code:




```{r}

set.seed(12345678)

N_fun <- function(X) sum(X)

D_fun <- function(X){
  d <- 0
  d <- d + sum(abs(X[1:3, ] - X[2:4, ]))
  d <- d + sum(abs(X[ ,1:3] - X[ ,2:4]))
  d
}
logpi <- function(X, p=0.8, lambda=0.5){
  n <- N_fun(X)
  d <- D_fun(X)
  n*log(p) + (16-n)*log(1-p) - lambda*d
}
diag_all1 <- function(X) all(diag(X) == 1)
mh_Q1_once <- function(n_iter=60000, burn=10000, p=0.8, lambda=0.5, X0=NULL,
                       keep_X = FALSE){
  if(is.null(X0)) X <- matrix(rbinom(16,1,0.5), 4,4) else X <- X0
  lp <- logpi(X,p,lambda)

  ind  <- logical(n_iter)
  Dtr  <- numeric(n_iter)
  acc  <- logical(n_iter)

  Xpost <- if(keep_X) matrix(NA_integer_, nrow = n_iter - burn, ncol = 16) else NULL
  k <- 0

  for(t in 1:n_iter){
    Y   <- matrix(rbinom(16,1,0.5), 4,4)
    lpy <- logpi(Y,p,lambda)

    if(log(runif(1)) < (lpy - lp)){
      X <- Y; lp <- lpy
      acc[t] <- TRUE
    }

    ind[t] <- diag_all1(X)
    Dtr[t] <- D_fun(X)

    if(keep_X && t > burn){
      k <- k + 1
      Xpost[k,] <- as.integer(c(X))
    }
  }

  list(
    indicator = ind[(burn+1):n_iter],
    D_trace   = Dtr[(burn+1):n_iter],
    acc_rate  = mean(acc),
    n_iter    = n_iter,
    burn      = burn,
    p         = p,
    lambda    = lambda,
    chain_last = X,
    X_post_vec = Xpost
  )
}
mh_Q2_once <- function(n_iter=60000, burn=10000, p=0.8, lambda=0.5, X0=NULL,
                       keep_X = FALSE){
  if(is.null(X0)) X <- matrix(rbinom(16,1,0.5), 4,4) else X <- X0
  lp <- logpi(X,p,lambda)

  ind  <- logical(n_iter)
  Dtr  <- numeric(n_iter)
  acc  <- logical(n_iter)

  Xpost <- if(keep_X) matrix(NA_integer_, nrow = n_iter - burn, ncol = 16) else NULL
  k <- 0
  for(t in 1:n_iter){
    Y <- X
    j <- sample.int(16, 1)
    Y[j] <- rbinom(1,1,0.5)
    lpy <- logpi(Y,p,lambda)

    if(log(runif(1)) < (lpy - lp)){
      X <- Y; lp <- lpy
      acc[t] <- TRUE
    }

    ind[t] <- diag_all1(X)
    Dtr[t] <- D_fun(X)

    if(keep_X && t > burn){
      k <- k + 1
      Xpost[k,] <- as.integer(c(X))
    }
  }

  list(
    indicator = ind[(burn+1):n_iter],
    D_trace   = Dtr[(burn+1):n_iter],
    acc_rate  = mean(acc),
    n_iter    = n_iter,
    burn      = burn,
    p         = p,
    lambda    = lambda,
    chain_last = X,
    X_post_vec = Xpost
  )
}


library(coda)

batch_se <- function(x, b = 200){
  x <- as.numeric(x)
  n <- length(x)
  a <- floor(n / b)
  x <- x[1:(a*b)]
  m <- matrix(x, nrow = b)
  bm <- colMeans(m)
  sqrt(var(bm) / a)
}

ess <- function(x) as.numeric(effectiveSize(as.numeric(x)))

make_plots <- function(res_Q1, res_Q2, main_tag=""){
  ind1 <- as.numeric(res_Q1$indicator)
  ind2 <- as.numeric(res_Q2$indicator)
  D1   <- res_Q1$D_trace
  D2   <- res_Q2$D_trace

  op <- par(no.readonly=TRUE); on.exit(par(op), add=TRUE)
  par(mfrow=c(2,3), mar=c(4,4,2,1))

  plot(ind1, type="l", xlab="iter (post-burn)", ylab="I(diag all 1)",
       main=paste0("Trace I: Q1 ", main_tag))
  plot(ind2, type="l", xlab="iter (post-burn)", ylab="I(diag all 1)",
       main=paste0("Trace I: Q2 ", main_tag))

  plot(D1, type="l", xlab="iter (post-burn)", ylab="D(X)",
       main=paste0("Trace D: Q1 ", main_tag))
  plot(D2, type="l", xlab="iter (post-burn)", ylab="D(X)",
       main=paste0("Trace D: Q2 ", main_tag))

  acf(ind1, main=paste0("ACF I: Q1 ", main_tag))
  acf(ind2, main=paste0("ACF I: Q2 ", main_tag))
}

summary_row <- function(res, algo="Q1", bsize=200){
  ind <- as.numeric(res$indicator)
  data.frame(
    lambda   = res$lambda,
    algo     = algo,
    n_post   = length(ind),
    acc_rate = res$acc_rate,
    est      = mean(ind),
    ESS      = ess(ind),
    SE_bm    = batch_se(ind, b=bsize),
    stringsAsFactors = FALSE
  )
}

p <- 0.8
lams <- c(0.5, 1.0)
n_iter <- 60000
burn   <- 10000
bsize  <- 200

OUT <- list(meta=list(p=p, lams=lams, n_iter=n_iter, burn=burn, bsize=bsize),
            runs=list())

for(lambda in lams){
  OUT$runs[[as.character(lambda)]] <- list(
    Q1 = mh_Q1_once(n_iter, burn, p, lambda, keep_X=FALSE),
    Q2 = mh_Q2_once(n_iter, burn, p, lambda, keep_X=FALSE)
  )
}


TAB <- do.call(rbind, lapply(lams, function(lambda){
  r <- OUT$runs[[as.character(lambda)]]
  rbind(summary_row(r$Q1, "Q1", bsize),
        summary_row(r$Q2, "Q2", bsize))
}))
#print(TAB)


#make_plots(OUT$runs[["0.5"]]$Q1, OUT$runs[["0.5"]]$Q2, main_tag="(lambda=0.5)")


#make_plots(OUT$runs[["1"]]$Q1, OUT$runs[["1"]]$Q2, main_tag="(lambda=1)")
```

We use the \textbf{Metropolis--Hastings (MH)} algorithm to approximate
$\Pr(\text{diag}(X) = \mathbf{1})$ under the distribution, as given in the question,
\[
\pi(X) \;\propto\;
p^{N(X)}\,(1-p)^{16-N(X)}\,\exp\!\bigl\{-\lambda\, D(X)\bigr\},
\]
with $p = 0.8$ and $\lambda \in \{0.5,\; 1.0\}$. We initialised the chain by drawing each of the 16 entries of the $4\times 4$ matrix $X^{(0)}$ independently from $\text{Bernoulli}(0.5)$. Each chain is run for $T = 60000$ iterations and we considered $10000$ as burn-in which we left out. After discarding the burnin-in, for the next $50000$ iterations we estimate the targer probability as, \[
    \widehat{P} \;=\; \frac{1}{T - T_{\text{burn}}}
    \sum_{t=T_{\text{burn}}+1}^{T} \mathbf{1}\!\bigl[\text{diag}(X^{(t)}) = \mathbf{1}\bigr].
    \]
Now for $Q1$, at each iteration the candidate $Y$ is generated by drawing all 16 entries independently from $\text{Bernoulli}(0.5)$, so we can write
\[
q_1(Y \mid X) \;=\; \prod_{i,j=1}^{4} \left(\tfrac{1}{2}\right) \;=\; \frac{1}{2^{16}}.
\]
 We set a seed at `12345678` for reproducibility. As becuase $q_1(Y\mid X) = q_1(X\mid Y)$, the proposal is symmetric we can actually simplify the MH acceptance probability to 
\[
\alpha(X, Y) \;=\; \min\!\left(1,\; \frac{\pi(Y)}{\pi(X)}\right).
\]
In our R code we worked on the log scale, so we accept $Y$ if $\log U < \log\pi(Y) - \log\pi(X)$, such that $U \sim \text{Uniform}(0,1)$.

Now for $Q2$, we have at each iterations one of the 16 entries being selected uniformly at random (with probability $1/16$) and set it to 0 or 1 with equal probability $1/2$, leaving all other entries totally unchanged. Foe example, if entry $(i,j)$ is chosen,
\[
Y_{kl} =
\begin{cases}
\text{Bernoulli}(0.5), & (k,l) = (i,j),\\
X_{kl}, & (k,l) \neq (i,j).
\end{cases}
\]
We note that even this is a symmetric proposal as the probability of choosing entry $(i,j)$ and setting it to a particular value is the same in both directions, so the acceptance probability again can be reduced to
\[
\alpha(X, Y) \;=\; \min\!\left(1,\; \frac{\pi(Y)}{\pi(X)}\right).
\]

## 2) 

From the table below, it is evident that the second algorithm , namely $Q2$ has a much faster mixing Markov Chain. We see that for that, for 60000 iterations with 10000 burn-ins, the effective sample size for $Q1$ is 51.6 whereas for $Q2$ its 970.8 for $\lambda=0.5$ and for $\lambda = 1$ we wet the effective samples sizes for $Q1$ and $Q2$ as 3.4 and 795.1 respective;y. The acceptance rate for $Q1$ is 0.008 and 0.0002 for $\lambda = 0.5,1$ respectively. For $Q2$, on the other hand has a much much higher acceptance rate of $0.6$ and $0.5$ for $\lambda = 0.5,1$ respectively. This shows that $Q1$ almost never accepts, especially for $\lambda =1$ making the chain extremely difficult to move. Further proof can be obtained by looking at the MCMC standard errors which are lower for $Q2$ (0.012 and 0.009 for the two $\lambda$ values in the respective orders) than $Q1$ (0.029 and 0.022 for the two $\lambda$ values in the respective orders). 


```{r}
TAB
```
To make our point more convincing, we can look at the ACF plots below. The first two ACF plots are for $\lambda =0.5$, precisely, first one for $Q1$ and the second for $Q2$. The next two are for $\lambda =1$ in the same order. We see that $Q1$ for both lamda values have high autocorrelation with its past values whereas for $Q2$ it decreases with increasing lags, this shows that for $Q1$ the chains will get stuck and will decay much more than in $Q2$ making it way faster.  



```{r}
par(mfrow = c(2,2), mar = c(4,4,4,1))

acf(as.numeric(OUT$runs[["0.5"]]$Q1$indicator),
    main = "ACF: Q1 (lambda = 0.5)")

acf(as.numeric(OUT$runs[["0.5"]]$Q2$indicator),
    main = "ACF: Q2 (lambda = 0.5)")

acf(as.numeric(OUT$runs[["1"]]$Q1$indicator),
    main = "ACF: Q1 (lambda = 1)")

acf(as.numeric(OUT$runs[["1"]]$Q2$indicator),
    main = "ACF: Q2 (lambda = 1)")
```


## 3) 

Here I am sharing all the 12 plots (6 for $\lambda=0.5$ and 6 for $\lambda=1$) needed to answer this question. For $\lambda =1$, we see that under $Q1$, the trace plot of $I_t$ shows long flat line at 1, showing that the chain becomes stuck in high-probability configurations. This is consistent with the extremely low acceptance rate (0.0002) and very very small ESS (3.4). However for $q2$, we can see that the trace plot shows continuous switching between 0 and 1, which proves that it is the space. Similar conclusion can be drawn for both for $\lambda=0.5$ For the trace of D(X), we see for $\lambda = 1$, $Q1$ quickly collapses  with very small D(X) and remains there whereas $Q2$ again explors a wider range. Similar observations can be seen for the case of $\lambda =0.5$ but with far less extreme behavior. For the autocorrelation plots, as explained before, $Q14$ shows high autocorrelation with its lagged values than $Q2$ indicating higher changes of chain being stuck. To decide on the effective sample size, we looked at the effective sample size, the 	Batch-means MCMC standard errors which we recorded in the next subpart. All these indicates that $Q2$ is a much much faster mixing MC than $Q1$, thus we are suured that $Q1$ will converge way faster than $Q1$. Here comes a trade off, running too many iterations will be useless for $Q2$ as it has possible converged way earlier and will lead to unnecessary computation time. On the other hand, if we stop as soon as $Q2$ converged, there will be a good chance that $Q1$ haven't converged at all. So a trade of can be made by looking at the mean plots (estimate vs. iteration). There if we see that $Q1$ is about to or just (kind of) convered to almost fixed estimated values and $Q2$ converges already before a few iteration, that would, we believe, an excellent place to stop. By looking at the running mean plots below, it looks like at 50000 iterations after a burn-in of 10000 looks exactly what we wanted, $Q1$ is alomst stable at running mean of $0.87$ and $Q2$ have already stabilized at $0.9$.



```{r}
make_plots(OUT$runs[["0.5"]]$Q1, OUT$runs[["0.5"]]$Q2, main_tag="(lambda=0.5)")
make_plots(OUT$runs[["1"]]$Q1, OUT$runs[["1"]]$Q2, main_tag="(lambda=1)")
```
```{r}
running_mean <- function(x){
  x <- as.numeric(x)
  cumsum(x) / seq_along(x)
}

running_mean_plots <- function(res_Q1, res_Q2, lambda){

  ind1 <- as.numeric(res_Q1$indicator)
  ind2 <- as.numeric(res_Q2$indicator)

  rm1 <- running_mean(ind1)
  rm2 <- running_mean(ind2)

  par(mfrow=c(1,2), mar=c(4,4,4,1))

  plot(rm1, type="l",
       xlab="Iteration (post-burn)",
       ylab="Running mean",
       main=paste0("Running mean: Q1 (lambda=", lambda, ")"))
  abline(h=mean(ind1), lty=3)

  plot(rm2, type="l",
       xlab="Iteration (post-burn)",
       ylab="Running mean",
       main=paste0("Running mean: Q2 (lambda=", lambda, ")"))
  abline(h=mean(ind2), lty=3)
}

# Run for both lambdas
running_mean_plots(OUT$runs[["0.5"]]$Q1,
                   OUT$runs[["0.5"]]$Q2,
                   0.5)

running_mean_plots(OUT$runs[["1"]]$Q1,
                   OUT$runs[["1"]]$Q2,
                   1)
```


## 4)

The following Table reports all estimates, along with associated MCMC standard errors (using consistent mini batches). 

```{r}
TAB
```

From the table above me see that, $Q1$ has 0.4619 and 0.8483 as estimates and 0.0296 and 0.02270 as standard errors for $\lambda$ at 0.5 and 1 respectively. Similarly $Q3$ has 0.7707 and 0.9190 as estimates and 0.01265 and 0.00918 as standard errors for $\lambda$ at 0.5 and 1 respectively.


## 5)

In Lecture 10, we have Monte Carlo Methods, there we are interested in estimating expectations of the form

$$\mu_g = \mathbb{E}_{\pi}[g(X)] = \int g(x)\,\pi(x)\,dx,$$

where $\pi$ is the target distribution. There we assumed that we can simulate draws from $\pi$. After that we can use Monte Carlo estimator $$\hat{\mu}_n = \frac{1}{n} \sum_{i=1}^n g(X_i),
\quad X_i \stackrel{iid}{\sim} \pi.$$

Now, we already know from Strong Law of Large Numbers, if $E_\pi|g(X)| < \infty$, then
$\hat{\mu}_n \xrightarrow{a.s.} \mu_g.$ From this, we get in the slides, the Central Limit Theorem which says, if $\tau^2 = \mathrm{Var}_\pi(g(X)) < \infty,$ then $\sqrt{n}(\hat{\mu}_n - \mu_g) \xrightarrow{d} N(0,\tau^2).$ This justifies the use of Monte Carlo estimation in the iid setting. However, in this problem we cannot directly sample iid from $\pi$ so we cannot actually get $\hat \mu_n$. To find a solution for this problem, we use Markov Chain with with Metropolis-Hasting algorithm with $\pi$ being the stationary distribution. There, as also described in slides, we construct Markov chain of $\pi$ as its stationary distribution, and take $n$ chain states as though they are draws from$ \pi$. Then we take averages again. We know, also written as a result in slides, that this method relies on Law of Large Numbers for Markov chains and the conditions are easily
satisfied. Thus, now we can actually get $\hat \mu_n$. Thus from discussion in class and slides given to us for Lecture 10, we can justify the use of the Markov chains we constructed above for estimating expectations with respect to $\pi$. 



## 6)

Now we make an attempt to construct a more efficient Metropolis-Hastings algorithm than the ones we have used above. We know A Gibbs Sampler is a special case of Metropolis Hasting algorithm with acceptance probability 1 since the proposal equals the full conditional. So the the idea is to use a Gibbs sampler ($Q3$) which would randomly visit all 16 cells, this should be more useful than $Q1$ and $Q2$. Here we sample $X_j$ for each cell $j$ from its full conditional distribution under $\pi$. 


$$
P(X_j = 1 \mid X_{-j}) = \frac{\pi(X \text{ with } X_j=1)}{\pi(X \text{ with } X_j=1) + \pi(X \text{ with } X_j=0)}
$$

Here we compute $\log\pi$ for both cases and use the softmax.



The following code implemends the Gibbs sampler and prepares a table that compares acceptance rate, batch Standard Error, etc. for the three. Here we compute $\log\pi$ for both cases ($P(X_j = 1 \mid X_{-j})$ and $P(X_j = 0 \mid X_{-j})$) and use the softmax operator.  As given in question, we only consider the case when $\lambda = 1.$





```{r}
set.seed(12345678)
N_fun <- function(X) sum(X)
D_fun <- function(X){
  sum(abs(X[1:3, ] - X[2:4, ])) + sum(abs(X[ ,1:3] - X[ ,2:4]))
}
logpi <- function(X, p=0.8, lambda=0.5){
  n <- N_fun(X)
  d <- D_fun(X)
  n*log(p) + (16-n)*log(1-p) - lambda*d
}
diag_all1 <- function(X) all(diag(X) == 1)
gibbs_once <- function(n_iter=60000, burn=10000, p=0.8, lambda=0.5, X0=NULL){
  if(is.null(X0)) X <- matrix(rbinom(16,1,0.5), 4,4) else X <- X0

  ind <- logical(n_iter)
  Dtr <- numeric(n_iter)

  for(t in 1:n_iter){
    perm <- sample.int(16)
    for(j in perm){
      X1 <- X; X1[j] <- 1
      X0tmp <- X; X0tmp[j] <- 0

      lp1 <- logpi(X1, p, lambda)
      lp0 <- logpi(X0tmp, p, lambda)

      prob1 <- 1 / (1 + exp(lp0 - lp1))
      X[j] <- rbinom(1, 1, prob1)
    }
    ind[t] <- diag_all1(X)
    Dtr[t] <- D_fun(X)
  }

  list(
    indicator  = ind[(burn+1):n_iter],
    D_trace    = Dtr[(burn+1):n_iter],
    acc_rate   = 1.0,
    n_iter     = n_iter,
    burn       = burn,
    p          = p,
    lambda     = lambda,
    chain_last = X
  )
}
mh_Q1_once <- function(n_iter=60000, burn=10000, p=0.8, lambda=0.5, X0=NULL){
  if(is.null(X0)) X <- matrix(rbinom(16,1,0.5), 4,4) else X <- X0
  lp <- logpi(X,p,lambda)

  ind <- logical(n_iter)
  Dtr <- numeric(n_iter)
  acc <- logical(n_iter)

  for(t in 1:n_iter){
    Y   <- matrix(rbinom(16,1,0.5), 4,4)
    lpy <- logpi(Y,p,lambda)
    if(log(runif(1)) < (lpy - lp)){
      X <- Y; lp <- lpy; acc[t] <- TRUE
    }
    ind[t] <- diag_all1(X)
    Dtr[t] <- D_fun(X)
  }

  list(
    indicator  = ind[(burn+1):n_iter],
    D_trace    = Dtr[(burn+1):n_iter],
    acc_rate   = mean(acc),
    n_iter     = n_iter,
    burn       = burn,
    p          = p,
    lambda     = lambda,
    chain_last = X
  )
}
mh_Q2_once <- function(n_iter=60000, burn=10000, p=0.8, lambda=0.5, X0=NULL){
  if(is.null(X0)) X <- matrix(rbinom(16,1,0.5), 4,4) else X <- X0
  lp <- logpi(X,p,lambda)

  ind <- logical(n_iter)
  Dtr <- numeric(n_iter)
  acc <- logical(n_iter)

  for(t in 1:n_iter){
    Y <- X
    j <- sample.int(16, 1)
    Y[j] <- rbinom(1,1,0.5)

    lpy <- logpi(Y,p,lambda)
    if(log(runif(1)) < (lpy - lp)){
      X <- Y; lp <- lpy; acc[t] <- TRUE
    }
    ind[t] <- diag_all1(X)
    Dtr[t] <- D_fun(X)
  }

  list(
    indicator  = ind[(burn+1):n_iter],
    D_trace    = Dtr[(burn+1):n_iter],
    acc_rate   = mean(acc),
    n_iter     = n_iter,
    burn       = burn,
    p          = p,
    lambda     = lambda,
    chain_last = X
  )
}
batch_se <- function(x, b = 200){
  x <- as.numeric(x)
  a <- floor(length(x) / b)
  x <- x[1:(a*b)]
  m <- matrix(x, nrow = b)
  bm <- colMeans(m)
  sqrt(var(bm) / a)
}
ess <- function(x) as.numeric(effectiveSize(as.numeric(x)))
summary_row <- function(res, algo="Q1", bsize=200){
  ind <- as.numeric(res$indicator)
  data.frame(
    lambda   = res$lambda,
    algo     = algo,
    n_post   = length(ind),
    acc_rate = round(res$acc_rate, 6),
    est      = round(mean(ind), 6),
    ESS      = round(ess(ind), 2),
    SE_bm    = round(batch_se(ind, b=bsize), 6),
    check.names = FALSE
  )
}
lambda <- 1.0
p <- 0.8
n_iter <- 60000
burn <- 10000
res_Q1 <- mh_Q1_once(n_iter, burn, p, lambda)
res_Q2 <- mh_Q2_once(n_iter, burn, p, lambda)
res_Q3 <- gibbs_once(n_iter, burn, p, lambda)
TAB <- rbind(
  summary_row(res_Q1, "Q1"),
  summary_row(res_Q2, "Q2"),
  summary_row(res_Q3, "Gibbs")
)
TAB
```


We see from the above table that, naturally Gibbs sample has acceptance probability 1. We see that the batch SE is much lower for the Gibbs ($Q3$) than for $Q1$ and $Q2$. Also effective sample size (ESS) is way larger for $Q3$ than $Q1$ and $Q2$. These proves $Q3$ having much faster mixing MC and is way efficient. 

```{r}
par(mfrow=c(3,2), mar=c(4,4,3,1))

plot(as.numeric(res_Q1$indicator), type="l",
     xlab="iter (post-burn)", ylab="I(diag all 1)",
     main="Trace I: Q1 (lambda=1)")
acf(as.numeric(res_Q1$indicator), main="ACF I: Q1 (lambda=1)")

plot(as.numeric(res_Q2$indicator), type="l",
     xlab="iter (post-burn)", ylab="I(diag all 1)",
     main="Trace I: Q2 (lambda=1)")
acf(as.numeric(res_Q2$indicator), main="ACF I: Q2 (lambda=1)")

plot(as.numeric(res_Q3$indicator), type="l",
     xlab="iter (post-burn)", ylab="I(diag all 1)",
     main="Trace I: Gibbs (lambda=1)")
acf(as.numeric(res_Q3$indicator), main="ACF I: Gibbs (lambda=1)")

par(mfrow=c(1,1))
```

We see from the above plopts that the Gibbs sampler's Trace plot is completely back showing that it covers the entire state space (that was natural from the construction). Also, it is no autocorrelation with its lags which proves that the chain would almost never get stuck at a state for the Gibbs sampler. On the contrary for $Q1$ and $Q2$ there are considerable autocorrelation with its lags which can prove chain being stuck for a prolonged time. Thus, from the table and from these plots, we can say that our proposed algorithm $Q3$ is way more efficient than $Q1$ and $Q2$. 


# Problem 5

We fit the nonlinear regression model:

$$Y_i = \frac{1 + \beta_1 X_i}{1 + \beta_2 \exp(\beta_3 X_i)} + \epsilon_i, \quad \epsilon_i \overset{iid}{\sim} N(0, \sigma^2)$$

with priors $\beta_j \sim N(0, \tau^2 = 10)$ for $j = 1,2,3$ and $\sigma^2 \sim \text{Gamma}(k=1, \theta=10)$ (shape-scale parameterization).

First we load the  Dataset with a very brief visualisation

```{r load-data}
library(MASS)
load("glm dat.RData")
n <- length(Y)
cat("n =", n, "\n")
plot(X, Y, pch = 16, col = rgb(0.27, 0.51, 0.71, 0.5),
     main = "Data: Y vs X", xlab = "X", ylab = "Y")
```

## 1)

We use **Metropolis-within-Gibbs** with an **adaptive block proposal** for the three $\beta$'s and a separate Metropolis step for $\sigma^2$ on the log-scale. The following is the pseudocode for this problem:

```
Initialize: beta = (3, 0.1, 0.3), sigma2 = 8
Set initial proposal covariance Sigma_prop = small diagonal matrix

For i = 2, ..., N_iter:

    # ── Adapt proposal (during warm-up only) ──
    If i <= adapt_end and i mod 500 == 0:
        Sigma_prop = (2.38^2 / 3) * Cov(recent samples) + eps * I

    # ── Block update (beta1, beta2, beta3) ──
    Propose beta* = beta_current + MVN(0, Sigma_prop)
    log_r = log_posterior(beta*, sigma2) - log_posterior(beta_curr, sigma2)
    Accept beta* with probability min(1, exp(log_r))

    # ── Update sigma^2 on log-scale ──
    Propose log(sigma2*) = log(sigma2_current) + Normal(0, sd_prop)
    log_r = log_post(beta, sigma2*) - log_post(beta, sigma2_curr)
            + log(sigma2*) - log(sigma2_curr)     [Jacobian]
    Accept sigma2* with probability min(1, exp(log_r))

    Store (beta1, beta2, beta3, sigma2)

Post-process: discard burn-in, thin every 5th sample
```

In framing this model, we made the following choices

- **Block proposal** for the three $\beta$'s because they are highly correlated in the posterior.
- **Adaptive covariance** using the $2.38^2/d$ scaling rule (Roberts & Rosenthal), frozen after adaptation.
- **Log-transform** for $\sigma^2$ to ensure positivity, with Jacobian correction.

Now we run the MCMC code:

```{r mcmc, cache=TRUE}
mu_func <- function(X, b1, b2, b3) {
  (1 + b1 * X) / (1 + b2 * exp(b3 * X))
}
log_posterior <- function(b1, b2, b3, sig2, X, Y, tau2 = 10) {
  mu <- mu_func(X, b1, b2, b3)
  if (any(!is.finite(mu))) return(-Inf)
  if (sig2 <= 0) return(-Inf)
  ll <- -n/2 * log(sig2) - sum((Y - mu)^2) / (2 * sig2)
  lp_betas <- -b1^2/(2*tau2) - b2^2/(2*tau2) - b3^2/(2*tau2)
  lp_sig2 <- -sig2 / 10
  return(ll + lp_betas + lp_sig2)
}
set.seed(42)
n_iter    <- 100000
burn_in   <- 20000
thin      <- 5
adapt_end <- 5000

samples <- matrix(NA, nrow = n_iter, ncol = 4)
colnames(samples) <- c("beta1", "beta2", "beta3", "sigma2")
samples[1, ] <- c(3, 0.1, 0.3, 8)

prop_cov_beta   <- diag(c(0.01, 0.005, 0.005))
prop_sd_logsig2 <- 0.15
accept_beta <- 0
accept_sig2 <- 0
for (i in 2:n_iter) {
  b_curr    <- samples[i-1, 1:3]
  sig2_curr <- samples[i-1, 4]
  
  
  if (i > 500 && i <= adapt_end && i %% 500 == 0) {
    emp_cov <- cov(samples[max(1, i-1000):(i-1), 1:3])
    if (all(is.finite(emp_cov)))
      prop_cov_beta <- 2.38^2 / 3 * emp_cov + 1e-6 * diag(3)
  }
  if (i == adapt_end + 1) {
    emp_cov <- cov(samples[1:(i-1), 1:3])
    prop_cov_beta <- 2.38^2 / 3 * emp_cov + 1e-6 * diag(3)
  }
  
  b_prop <- b_curr + mvrnorm(1, mu = rep(0, 3), Sigma = prop_cov_beta)
  lp_curr <- log_posterior(b_curr[1], b_curr[2], b_curr[3], sig2_curr, X, Y)
  lp_prop <- log_posterior(b_prop[1], b_prop[2], b_prop[3], sig2_curr, X, Y)
  log_r <- lp_prop - lp_curr
  if (is.finite(log_r) && log(runif(1)) < log_r) {
    b_curr <- b_prop
    accept_beta <- accept_beta + 1
  }
  
  log_sig2_curr <- log(sig2_curr)
  log_sig2_prop <- rnorm(1, log_sig2_curr, prop_sd_logsig2)
  sig2_prop <- exp(log_sig2_prop)
  lp_curr_s <- log_posterior(b_curr[1], b_curr[2], b_curr[3], sig2_curr, X, Y)
  lp_prop_s <- log_posterior(b_curr[1], b_curr[2], b_curr[3], sig2_prop, X, Y)
  log_r_s <- lp_prop_s - lp_curr_s + log_sig2_prop - log_sig2_curr
  if (is.finite(log_r_s) && log(runif(1)) < log_r_s) {
    sig2_curr <- sig2_prop
    accept_sig2 <- accept_sig2 + 1
  }
  
  samples[i, ] <- c(b_curr, sig2_curr)
}

keep <- seq(burn_in + 1, n_iter, by = thin)
post <- samples[keep, ]
```

Upon running the MCMC we get the following acceptance rates:

```{r acceptance}
cat("Betas (block):", round(accept_beta / n_iter * 100, 1), "%\n")
cat("Sigma^2:      ", round(accept_sig2 / n_iter * 100, 1), "%\n")
cat("Post burn-in/thin samples:", nrow(post), "\n")
```

## 2)


```{r trace-plots, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (j in 1:4) {
  plot(post[, j], type = "l", col = "steelblue",
       ylab = colnames(post)[j], xlab = "Iteration (post burn-in)",
       main = paste("Trace:", colnames(post)[j]))
}
```

All four trace plots look stationary --- the chains bounce around a stable mean with no obvious trends or drift, indicating convergence. This proves that the burn-in of 20,000 iterations were sufficient.

For further conformation, we can have a look at the acf plots of the coefficients:

```{r acf-plots, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (j in 1:4) {
  acf(post[, j], main = paste("ACF:", colnames(post)[j]), lag.max = 50)
}
```

For the three $\beta$'s, the autocorrelation starts high (~0.8) and decays to near zero by about lag 20--30. For $\sigma^2$, the ACF drops below the significance band almost immediately (by lag 2--3), showing its excellent mixing.

Next, we have a look at the effective sample sizes (ESS) for the coefficients:

```{r ess}
ess_func <- function(x) {
  nn <- length(x)
  acf_vals <- acf(x, plot = FALSE, lag.max = min(500, nn-1))$acf[-1]
  cutoff <- which(acf_vals < 0.05)[1]
  if (is.na(cutoff)) cutoff <- length(acf_vals)
  tau <- 1 + 2 * sum(acf_vals[1:cutoff])
  return(nn / max(tau, 1))
}

ess_vals <- sapply(1:4, function(j) round(ess_func(post[, j])))
names(ess_vals) <- colnames(post)
print(ess_vals)
```

All ESS values are above 1,000 for the $\beta$'s and above 11,000 for $\sigma^2$, this again hints at that the MCMC approximations are reliable.

## 3)


In this sub part we summarize approximations of the posterior expectations for each of the 4 variables.

```{r summaries}
post_means <- colMeans(post)

mcmc_se <- function(x, n_batches = 50) {
  nn <- length(x)
  bs <- floor(nn / n_batches)
  bm <- sapply(1:n_batches, function(b) mean(x[((b-1)*bs+1):(b*bs)]))
  sd(bm) / sqrt(n_batches)
}

results <- data.frame(
  Parameter = colnames(post),
  Post_Mean = round(post_means, 4),
  MCMC_SE   = sapply(1:4, function(j) round(mcmc_se(post[, j]), 6)),
  CI_Lower  = round(apply(post, 2, quantile, 0.025), 4),
  CI_Upper  = round(apply(post, 2, quantile, 0.975), 4)
)
knitr::kable(results, col.names = c("Parameter", "Post. Mean", "MCMC SE",
                                     "95% CI Lower", "95% CI Upper"),
             row.names = FALSE)
```

The MCMC standard errors are very small relative to the posterior standard deviations. This shows that the chain ran long enough and we have obtained reliable estimation.

## 4) 

Now we plot the posterior pdfs for the for coefficients:

```{r posterior-pdfs, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (j in 1:4) {
  d <- density(post[, j])
  plot(d, main = paste("Posterior:", colnames(post)[j]),
       xlab = colnames(post)[j], col = "steelblue", lwd = 2)
  polygon(d, col = rgb(0.27, 0.51, 0.71, 0.3), border = "steelblue")
  abline(v = post_means[j], col = "red", lty = 2, lwd = 1.5)
  legend("topright", legend = sprintf("Mean = %.3f", post_means[j]),
         col = "red", lty = 2, bty = "n", cex = 0.9)
}
```

All four posteriors are roughly unimodal. The $\beta$'s show some mild skewness, while $\sigma^2$ has a slight right skew pattern.

## 5) 

Now we  investigate Posterior Dependence, first through correlation matrix.


```{r correlations}
cor_mat <- cor(post)
knitr::kable(round(cor_mat, 3), caption = "Posterior Correlation Matrix")
```

From above we can see that $\beta_1$, $\beta_2$ and $\beta_3$ are pretty strongly correlated. However $\sigma^2$ seemed independent of the $\beta$ parameters. 

Next, as asked in the question, we show Bivariate Scatter Plots:

```{r bivariate, fig.height=8}
pairs(post, col = rgb(0.27, 0.51, 0.71, 0.1), pch = 16, cex = 0.5,
      main = "Bivariate Posterior Samples", labels = colnames(post))
```


We again note that there is **very strong positive correlation** among all three $\beta$ parameters.


Hence we are double confirmed about the high correlations which means the three parameters are **not easily identifiable separately** --- a change in one $\beta$ can be compensated by changes in the others to produce a similar fit. The bivariate scatter plots show elongated elliptical clouds, confirming this. This is why we needed the block proposal: updating $\beta$'s one at a time would move along the narrow direction of these ellipses, leading to very slow mixing.

As just noted before, $\sigma^2$ is essentially independent of the three $\beta$'s ($|r| < 0.025$ for all pairs), which makes sense to us as the error variance is orthogonal to the mean function parameters once the residuals are computed.

## 6)

No we implemend Importance Sampling with prior Variance of 5. We reweight the MCMC samples (obtained under $\tau^2 = 10$) to approximate the posterior under $\tau^2 = 5$. The IS weight for each sample is:

$$w^{(i)} \propto \prod_{j=1}^{3} \frac{N(\beta_j^{(i)}; 0, 5)}{N(\beta_j^{(i)}; 0, 10)}$$

```{r is-5}
log_is_w5 <- apply(post[, 1:3], 1, function(b) sum(0.5*log(2) - b^2/20))
log_w5 <- log_is_w5 - max(log_is_w5)
w5 <- exp(log_w5)
w5n <- w5 / sum(w5)
ess5 <- 1 / sum(w5n^2)

is_results_5 <- data.frame(
  Parameter = colnames(post),
  IS_Mean = sapply(1:4, function(j) round(sum(w5n * post[, j]), 4)),
  IS_SE   = sapply(1:4, function(j) {
    m <- sum(w5n * post[, j])
    v <- sum(w5n * (post[, j] - m)^2)
    round(sqrt(v / ess5), 5)
  })
)
knitr::kable(is_results_5, col.names = c("Parameter", "IS Mean (tau2=5)", "IS SE"),
             row.names = FALSE)
cat("IS effective sample size:", round(ess5, 1), "\n")
```

The IS ESS is nearly the full sample size (15954.5 out of 16000), indicating excellent weight stability. The tighter prior ($\tau^2 = 5$) places more mass where our samples already concentrate, so essentially all samples remain informative. 

## 7) 

Now we reimplemend Importance Sampling with Prior Variance of 50.

```{r is-50}
log_is_w50 <- apply(post[, 1:3], 1, function(b) sum(-0.5*log(5) + b^2/25))
log_w50 <- log_is_w50 - max(log_is_w50)
w50 <- exp(log_w50)
w50n <- w50 / sum(w50)
ess50 <- 1 / sum(w50n^2)

is_results_50 <- data.frame(
  Parameter = colnames(post),
  IS_Mean = sapply(1:4, function(j) round(sum(w50n * post[, j]), 4)),
  IS_SE   = sapply(1:4, function(j) {
    m <- sum(w50n * post[, j])
    v <- sum(w50n * (post[, j] - m)^2)
    round(sqrt(v / ess50), 5)
  })
)
knitr::kable(is_results_50, col.names = c("Parameter", "IS Mean (tau2=50)", "IS SE"),
             row.names = FALSE)
cat("IS effective sample size:", round(ess50, 1), "\n")
```

The IS ESS is again very very good, and the posterior means pretty much did not change across all three prior specifications ($\tau^2 = 5, 10, 50$). With $n = 200$ observations, the data wins over the prior andwhether we use $\tau^2 = 5$ or 50 it is making very little of difference to the posterior. This is reassuring as our inference is robust to the choice of prior variance for the $\beta$'s.


