---
title: "HW2"
author: "Mengyan Jing and Abhishek Banerjee"
date: "2026-02-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1


```{r}
integrand <- function(t, s) {
  # Use log-scale computation to avoid overflow/underflow
  exp((s - 1) * log(t) - t)
}

lower_inc_gamma <- function(s, x) {
  if (x <= 0) return(0)
  
  # For large x, compute as gamma(s) - upper
  if (x > 700) {
    return(gamma(s) - upper_inc_gamma(s, x))
  }
  
  # Split: integrate from a small epsilon to x
  # Near 0, use substitution u = t^s to remove singularity
  # Original: integral of t^(s-1)*exp(-t) dt from 0 to x
  # Sub u = t^s => t = u^(1/s), dt = (1/s)*u^(1/s - 1) du
  # Integral becomes (1/s) * integral of exp(-u^(1/s)) du from 0 to x^s
  
  if (s < 1) {
    # Use substitution
    transformed_integrand <- function(u, s) {
      (1/s) * exp(-u^(1/s))
    }
    result <- integrate(transformed_integrand, lower = 0, upper = x^s, s = s,
                        rel.tol = 1e-10, subdivisions = 1000L)
  } else {
    result <- integrate(integrand, lower = 0, upper = x, s = s,
                        rel.tol = 1e-10, subdivisions = 1000L)
  }
  return(result$value)
}

upper_inc_gamma <- function(s, x) {
  if (x <= 0) return(gamma(s))
  result <- integrate(integrand, lower = x, upper = Inf, s = s,
                      rel.tol = 1e-10, subdivisions = 1000L)
  return(result$value)
}

lower_inc_gamma_vec <- function(s_vec, x_vec) {
  mapply(lower_inc_gamma, s_vec, x_vec)
}

upper_inc_gamma_vec <- function(s_vec, x_vec) {
  mapply(upper_inc_gamma, s_vec, x_vec)
}
```

```{r}
s_vals <- seq(0.0001, 10, length.out = 100)
x_vals <- seq(0.0001, 10000, length.out = 100)
param_set <- expand.grid(s = s_vals, x = x_vals)

# Time lower
ptm <- proc.time()
lower_results <- lower_inc_gamma_vec(param_set$s, param_set$x)
proc.time() - ptm

# Time upper
ptm <- proc.time()
upper_results <- upper_inc_gamma_vec(param_set$s, param_set$x)
proc.time() - ptm

```


```{r}
# Baseline from R's built-in functions
baseline_lower <- pgamma(param_set$x, shape = param_set$s, scale = 1, 
                          lower.tail = TRUE) * gamma(param_set$s)
baseline_upper <- pgamma(param_set$x, shape = param_set$s, scale = 1, 
                          lower.tail = FALSE) * gamma(param_set$s)

# MSE
mse_lower <- mean((lower_results - baseline_lower)^2)
mse_upper <- mean((upper_results - baseline_upper)^2)

cat("MSE lower:", mse_lower, "\n")
cat("MSE upper:", mse_upper, "\n")

```


```{r}
library(Rcpp)
```

```{r}
Sys.which("make")
```

## Task 1

```{Rcpp}
#include <Rcpp.h>
#include <cmath>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector integrand_cpp(NumericVector t, double s) {
  int n = t.size();
  NumericVector out(n);
  for(int i=0; i<n; i++){
    out[i] = std::exp((s - 1.0) * std::log(t[i]) - t[i]);
  }
  return out;
}

// [[Rcpp::export]]
NumericVector transformed_integrand_cpp(NumericVector u, double s) {
  int n = u.size();
  NumericVector out(n);
  for(int i=0; i<n; i++){
    out[i] = (1.0 / s) * std::exp(-std::pow(u[i], 1.0 / s));
  }
  return out;
}
```



```{r}
lower_inc_gamma_cpp <- function(s, x) {
  if (x <= 0) return(0)
  
  if (x > 700) {
    return(gamma(s) - upper_inc_gamma_cpp(s, x))
  }
  
  if (s < 1) {
    result <- integrate(function(u) transformed_integrand_cpp(u, s),
                        lower = 0, upper = x^s,
                        rel.tol = 1e-10, subdivisions = 1000L)
  } else {
    result <- integrate(function(t) integrand_cpp(t, s),
                        lower = 0, upper = x,
                        rel.tol = 1e-10, subdivisions = 1000L)
  }
  return(result$value)
}

upper_inc_gamma_cpp <- function(s, x) {
  if (x <= 0) return(gamma(s))
  result <- integrate(function(t) integrand_cpp(t, s),
                      lower = x, upper = Inf,
                      rel.tol = 1e-10, subdivisions = 1000L)
  return(result$value)
}

lower_inc_gamma_cpp_vec <- function(s_vec, x_vec) {
  mapply(lower_inc_gamma_cpp, s_vec, x_vec)
}

upper_inc_gamma_cpp_vec <- function(s_vec, x_vec) {
  mapply(upper_inc_gamma_cpp, s_vec, x_vec)
}
```


```{r}
# Timing
ptm <- proc.time()
lower_results_cpp <- lower_inc_gamma_cpp_vec(param_set$s, param_set$x)
proc.time() - ptm

ptm <- proc.time()
upper_results_cpp <- upper_inc_gamma_cpp_vec(param_set$s, param_set$x)
proc.time() - ptm

# MSE
mse_lower_cpp <- mean((lower_results_cpp - baseline_lower)^2)
mse_upper_cpp <- mean((upper_results_cpp - baseline_upper)^2)
cat("MSE lower (C++ integrand):", mse_lower_cpp, "\n")
cat("MSE upper (C++ integrand):", mse_upper_cpp, "\n")
```


## 3)

```{r}
# install.packages("statmod")
library(statmod)

gw <- gauss.quad(n = 30, kind = "legendre")
nodes <- gw$nodes
weights <- gw$weights

nodes
weights
```


```{Rcpp}
#include <Rcpp.h>
#include <cmath>
#include <vector>
using namespace Rcpp;

// 30-point Gauss-Legendre nodes and weights
void gauss_legendre_30(std::vector<double>& nodes, std::vector<double>& weights) {
  const double x[] = {
    0.0514718426, 0.1538699136, 0.2546369261, 0.3527047255, 0.4470337695,
    0.5366241481, 0.6205261830, 0.6978504948, 0.7677774321, 0.8295657624,
    0.8825605358, 0.9262000474, 0.9600218650, 0.9836681233, 0.9968934841
  };
  const double w[] = {
    0.1028526529, 0.1017623897, 0.0995934206, 0.0963687372, 0.0921225222,
    0.0868997872, 0.0807558952, 0.0737559747, 0.0659742299, 0.0574931562,
    0.0484026728, 0.0387991926, 0.0287847079, 0.0184664683, 0.0079681925
  };
  nodes.resize(30);
  weights.resize(30);
  for (int i = 0; i < 15; i++) {
    nodes[i]        = -x[14 - i];
    weights[i]      =  w[14 - i];
    nodes[15 + i]   =  x[i];
    weights[15 + i] =  w[i];
  }
}

static std::vector<double> gl_nodes, gl_wts;
static bool gl_init = false;

void ensure_gl() {
  if (!gl_init) {
    gauss_legendre_30(gl_nodes, gl_wts);
    gl_init = true;
  }
}

// Generic GL integration over [a,b] with n_panels
typedef double (*integrand_fn)(double, double, double);

double gl_integrate2(double a, double b, int n_panels,
                     integrand_fn f, double p1, double p2) {
  ensure_gl();
  double h = (b - a) / n_panels;
  double total = 0.0;
  for (int p = 0; p < n_panels; p++) {
    double mid = a + h * (p + 0.5);
    double half_h = h * 0.5;
    for (int i = 0; i < 30; i++) {
      double t = mid + half_h * gl_nodes[i];
      total += half_h * gl_wts[i] * f(t, p1, p2);
    }
  }
  return total;
}

// Direct integrand: t^(s-1)*exp(-t)
double direct_ig(double t, double s, double unused) {
  if (t <= 0.0) return 0.0;
  return std::exp((s - 1.0) * std::log(t) - t);
}

// Transformed for s<1: (1/s)*exp(-u^(1/s))
double transformed_ig(double u, double s, double unused) {
  if (u <= 0.0) return 0.0;
  return (1.0 / s) * std::exp(-std::pow(u, 1.0 / s));
}

// ✅ NEW: stable upper integrand on v ∈ [0, Vmax], with t = x + v
double upper_v_ig(double v, double s, double x) {
  double t = x + v;
  if (t <= 0.0) return 0.0;
  return std::exp((s - 1.0) * std::log(t) - t);
}

// Panels heuristic
int choose_panels(double range) {
  if (range < 1.0) return 20;
  if (range < 10.0) return 40;
  if (range < 100.0) return 80;
  if (range < 1000.0) return 160;
  return 300;
}

// ✅ Scalar helpers so upper can call lower (pure C++)
double lower_scalar(double s, double x) {
  if (x <= 0.0) return 0.0;

  if (s < 1.0) {
    double upper_lim = std::pow(x, s);
    int panels = choose_panels(upper_lim);
    return gl_integrate2(1e-15, upper_lim, panels, transformed_ig, s, 0.0);
  } else {
    int panels = choose_panels(x);
    return gl_integrate2(1e-15, x, panels, direct_ig, s, 0.0);
  }
}

double upper_scalar(double s, double x) {
  if (x <= 0.0) return R::gammafn(s);

  // very large x => ~0
  double cutoff = s + 50.0 * std::sqrt(s) + 200.0;
  if (x > cutoff) return 0.0;

  // ✅ Key fix: for x < s, use complement (avoids missing mass)
  if (x < s + 2.0 * std::sqrt(s)) {
    return R::gammafn(s) - lower_scalar(s, x);
  }

  // ✅ Otherwise compute upper directly, stable v-integral (no 1/u)
  double Vmax = 60.0 + 10.0 * std::sqrt(s);   // baseline
  if (x > 10.0) Vmax = 40.0 + 8.0 * std::sqrt(s);
  if (x > 50.0) Vmax = 25.0 + 5.0 * std::sqrt(s);   // safe truncation
  int panels = choose_panels(Vmax);
  return gl_integrate2(0.0, Vmax, panels, upper_v_ig, s, x);
}

// [[Rcpp::export]]
NumericVector lower_inc_gamma_full_cpp(NumericVector s_vec, NumericVector x_vec) {
  int n = s_vec.size();
  NumericVector out(n);

  for (int i = 0; i < n; i++) {
    double s = s_vec[i];
    double x = x_vec[i];

    if (x <= 0.0) { out[i] = 0.0; continue; }

    // For huge x, lower = gamma(s) - upper (upper tiny)
    if (x > 700.0) {
      out[i] = R::gammafn(s) - upper_scalar(s, x);
    } else {
      out[i] = lower_scalar(s, x);
    }
  }
  return out;
}

// [[Rcpp::export]]
NumericVector upper_inc_gamma_full_cpp(NumericVector s_vec, NumericVector x_vec) {
  int n = s_vec.size();
  NumericVector out(n);

  for (int i = 0; i < n; i++) {
    out[i] = upper_scalar((double)s_vec[i], (double)x_vec[i]);
  }
  return out;
}
```



```{r}
# Timing
ptm <- proc.time()
lower_results_full <- lower_inc_gamma_full_cpp(param_set$s, param_set$x)
proc.time() - ptm

ptm <- proc.time()
upper_results_full <- upper_inc_gamma_full_cpp(param_set$s, param_set$x)
proc.time() - ptm

# MSE
mse_lower_full <- mean((lower_results_full - baseline_lower)^2)
mse_upper_full <- mean((upper_results_full - baseline_upper)^2)
cat("MSE lower (full C++):", mse_lower_full, "\n")
cat("MSE upper (full C++):", mse_upper_full, "\n")
```

# Problem 3

We have in the question, an exponentially tilted positive stable distribution which is very useful and flexible in modeling extreme weather. The family of such distributions is parameterized by $\theta > 0$. We have its density function as:
$$
f(x;\theta)
=
\frac{\sqrt{\frac{1}{4\pi x^{3}}}\,
\exp\!\left(-\frac{1}{4x} - \theta x\right)}
{\exp(-\sqrt{\theta})},
\qquad x > 0.
$$
## 1)

We will use Rejection Sampling technique discueesed in class for this problem. If we look at the above density and if we ignore the constant terms, we get the kernal as $$x^{-3/2} \exp\!\left(-\frac{1}{4x} - \theta x\right)$$. This kernal, at least to us, looks strikingly similar to that of the kernel of an Inverse Gaussian distribution. From wikipedia, I got the density of Inverse Gaussian distribution as $$
f(x;\mu,\lambda)
=
\sqrt{\frac{\lambda}{2\pi x^{3}}}
\exp\!\left(
-\frac{\lambda (x-\mu)^2}{2\mu^{2}x}
\right),
\qquad x>0,$$ where $\mu >0$ and $\lambda > 0$ are parameters. Thus, we can use this distribution as a proposal distribution. We can use in-built R function to sample from this proposal distribution.  We use the following R code to implement the Rejection sampling with Inverse Gaussian being our proposal distribution. 

```{r}
library(statmod)   

# Target density
f_theta <- function(x, theta = 0.1){
  ifelse(x > 0,
         sqrt(1/(4*pi*x^3)) *
           exp(-1/(4*x) - theta*x + sqrt(theta)),
         0)
}

# Proposal: Inverse Gaussian
g_prop <- function(x, theta = 0.1){
  mu  <- 1/(2*sqrt(theta))
  lam <- 1
  dinvgauss(x, mean = mu, shape = lam)
}

# Rejection sampler 
tilted_sampler <- function(theta = 0.1){

  mu  <- 1/(2*sqrt(theta))
  lam <- 1

  counter <- 1

  x <- rinvgauss(1, mean = mu, shape = lam)
  ratio <- f_theta(x, theta) / g_prop(x, theta)

  while(runif(1) > ratio){

    x <- rinvgauss(1, mean = mu, shape = lam)
    counter <- counter + 1

    ratio <- f_theta(x, theta) / g_prop(x, theta)
  }

  return(c(x, counter))
}
```

Now as asked in the question, we sample $N=100000$ independent samples for $\theta= 0.1$


```{r}
N <- 100000
res <- sapply(1:N, function(i) tilted_sampler(0.1))

samples     <- res[1,]
num_sampled <- sum(res[2,])
```


## 2) 

Now we replicate Figure 3 in the question; that is, plot the sample density using histogram and compare it to the underlying true density. We use the ocde given in the question:

```{r}
hist(samples, breaks = 1000, freq = FALSE,
     border = "lightblue")

curve(f_theta(x, 0.1),
      from = 1e-5, to = 60,
      n = 1000, add = TRUE,
      col = "red", lty = 3)
```

We see in the above plot, there is a perfect alignment between the sample density and the analytic density. This showed that our choice of proposal distribution was correct and the rejection sampling technique henceforth did a good job!



# Problem 4 

In this question we consider a 4 × 4 grid of Bernoulli random variables denoted by $X = [Xij]$. We also consider $N(X)$ as the total number of successes in X (the sum of X) and $D(X)$ as the total number of neighbors (horizontal or vertical) which differ. We are also told that if we write the elements of X out in a vector V , then there is a 24×16 matrix M such that $D(X)$ is sum of the absolute values of MV . The 24 arises because there are 24 neighbor-pairs to check (diagonal elements are not neighbors.) We have probability distribution of X being proportional to $$
\pi(X) \propto 
p^{N(X)} (1 - p)^{16 - N(X)} 
\exp\!\left\{ -\lambda D(X) \right\}.$$
Therefore, if $\lambda = 0$, we have i.i.d.\ Bernoulli entries. We use Markov chain Monte Carlo to approximate the probability that the diagonal of this matrix is all 1’s. We assume $p = 0.8$. We approximate the probabilities for each of the following cases: $\lambda = 0.5, \: \lambda = 1.$ We have $Q1$ where every transition is equally likely with probability $1/2^{16}$. That is, the next candidate state is simply 16 i.i.d. Bernoulli variables with prob- ability 0.5 of success each. We use $Q2$ such that we choose one of the 16 entries in the table with probability 1/16 and then set that entry to 0 or 1 with probability 0.5 each. We use the following R code:




```{r}

set.seed(12345678)

N_fun <- function(X) sum(X)

D_fun <- function(X){
  d <- 0
  d <- d + sum(abs(X[1:3, ] - X[2:4, ]))
  d <- d + sum(abs(X[ ,1:3] - X[ ,2:4]))
  d
}
logpi <- function(X, p=0.8, lambda=0.5){
  n <- N_fun(X)
  d <- D_fun(X)
  n*log(p) + (16-n)*log(1-p) - lambda*d
}
diag_all1 <- function(X) all(diag(X) == 1)
mh_Q1_once <- function(n_iter=60000, burn=10000, p=0.8, lambda=0.5, X0=NULL,
                       keep_X = FALSE){
  if(is.null(X0)) X <- matrix(rbinom(16,1,0.5), 4,4) else X <- X0
  lp <- logpi(X,p,lambda)

  ind  <- logical(n_iter)
  Dtr  <- numeric(n_iter)
  acc  <- logical(n_iter)

  Xpost <- if(keep_X) matrix(NA_integer_, nrow = n_iter - burn, ncol = 16) else NULL
  k <- 0

  for(t in 1:n_iter){
    Y   <- matrix(rbinom(16,1,0.5), 4,4)
    lpy <- logpi(Y,p,lambda)

    if(log(runif(1)) < (lpy - lp)){
      X <- Y; lp <- lpy
      acc[t] <- TRUE
    }

    ind[t] <- diag_all1(X)
    Dtr[t] <- D_fun(X)

    if(keep_X && t > burn){
      k <- k + 1
      Xpost[k,] <- as.integer(c(X))
    }
  }

  list(
    indicator = ind[(burn+1):n_iter],
    D_trace   = Dtr[(burn+1):n_iter],
    acc_rate  = mean(acc),
    n_iter    = n_iter,
    burn      = burn,
    p         = p,
    lambda    = lambda,
    chain_last = X,
    X_post_vec = Xpost
  )
}
mh_Q2_once <- function(n_iter=60000, burn=10000, p=0.8, lambda=0.5, X0=NULL,
                       keep_X = FALSE){
  if(is.null(X0)) X <- matrix(rbinom(16,1,0.5), 4,4) else X <- X0
  lp <- logpi(X,p,lambda)

  ind  <- logical(n_iter)
  Dtr  <- numeric(n_iter)
  acc  <- logical(n_iter)

  Xpost <- if(keep_X) matrix(NA_integer_, nrow = n_iter - burn, ncol = 16) else NULL
  k <- 0
  for(t in 1:n_iter){
    Y <- X
    j <- sample.int(16, 1)
    Y[j] <- rbinom(1,1,0.5)
    lpy <- logpi(Y,p,lambda)

    if(log(runif(1)) < (lpy - lp)){
      X <- Y; lp <- lpy
      acc[t] <- TRUE
    }

    ind[t] <- diag_all1(X)
    Dtr[t] <- D_fun(X)

    if(keep_X && t > burn){
      k <- k + 1
      Xpost[k,] <- as.integer(c(X))
    }
  }

  list(
    indicator = ind[(burn+1):n_iter],
    D_trace   = Dtr[(burn+1):n_iter],
    acc_rate  = mean(acc),
    n_iter    = n_iter,
    burn      = burn,
    p         = p,
    lambda    = lambda,
    chain_last = X,
    X_post_vec = Xpost
  )
}


library(coda)

batch_se <- function(x, b = 200){
  x <- as.numeric(x)
  n <- length(x)
  a <- floor(n / b)
  x <- x[1:(a*b)]
  m <- matrix(x, nrow = b)
  bm <- colMeans(m)
  sqrt(var(bm) / a)
}

ess <- function(x) as.numeric(effectiveSize(as.numeric(x)))

make_plots <- function(res_Q1, res_Q2, main_tag=""){
  ind1 <- as.numeric(res_Q1$indicator)
  ind2 <- as.numeric(res_Q2$indicator)
  D1   <- res_Q1$D_trace
  D2   <- res_Q2$D_trace

  op <- par(no.readonly=TRUE); on.exit(par(op), add=TRUE)
  par(mfrow=c(2,3), mar=c(4,4,2,1))

  plot(ind1, type="l", xlab="iter (post-burn)", ylab="I(diag all 1)",
       main=paste0("Trace I: Q1 ", main_tag))
  plot(ind2, type="l", xlab="iter (post-burn)", ylab="I(diag all 1)",
       main=paste0("Trace I: Q2 ", main_tag))

  plot(D1, type="l", xlab="iter (post-burn)", ylab="D(X)",
       main=paste0("Trace D: Q1 ", main_tag))
  plot(D2, type="l", xlab="iter (post-burn)", ylab="D(X)",
       main=paste0("Trace D: Q2 ", main_tag))

  acf(ind1, main=paste0("ACF I: Q1 ", main_tag))
  acf(ind2, main=paste0("ACF I: Q2 ", main_tag))
}

summary_row <- function(res, algo="Q1", bsize=200){
  ind <- as.numeric(res$indicator)
  data.frame(
    lambda   = res$lambda,
    algo     = algo,
    n_post   = length(ind),
    acc_rate = res$acc_rate,
    est      = mean(ind),
    ESS      = ess(ind),
    SE_bm    = batch_se(ind, b=bsize),
    stringsAsFactors = FALSE
  )
}

p <- 0.8
lams <- c(0.5, 1.0)
n_iter <- 60000
burn   <- 10000
bsize  <- 200

OUT <- list(meta=list(p=p, lams=lams, n_iter=n_iter, burn=burn, bsize=bsize),
            runs=list())

for(lambda in lams){
  OUT$runs[[as.character(lambda)]] <- list(
    Q1 = mh_Q1_once(n_iter, burn, p, lambda, keep_X=FALSE),
    Q2 = mh_Q2_once(n_iter, burn, p, lambda, keep_X=FALSE)
  )
}

# ============================================================
# TABLE for Part (4): estimates + MCMC SE + ESS + acc rate
# ============================================================

TAB <- do.call(rbind, lapply(lams, function(lambda){
  r <- OUT$runs[[as.character(lambda)]]
  rbind(summary_row(r$Q1, "Q1", bsize),
        summary_row(r$Q2, "Q2", bsize))
}))
print(TAB)

# ============================================================
# FIGURES for Part (2)-(3): make representative plots
# (call these when writing up; chains already stored in OUT)
# ============================================================

# Example: plots for lambda=0.5
make_plots(OUT$runs[["0.5"]]$Q1, OUT$runs[["0.5"]]$Q2, main_tag="(lambda=0.5)")

# Example: plots for lambda=1
make_plots(OUT$runs[["1"]]$Q1, OUT$runs[["1"]]$Q2, main_tag="(lambda=1)")
```

We use the \textbf{Metropolis--Hastings (MH)} algorithm to approximate
$\Pr(\text{diag}(X) = \mathbf{1})$ under the distribution, as given in the question,
\[
\pi(X) \;\propto\;
p^{N(X)}\,(1-p)^{16-N(X)}\,\exp\!\bigl\{-\lambda\, D(X)\bigr\},
\]
with $p = 0.8$ and $\lambda \in \{0.5,\; 1.0\}$. We initialised the chain by drawing each of the 16 entries of the $4\times 4$ matrix $X^{(0)}$ independently from $\text{Bernoulli}(0.5)$. Each chain is run for $T = 60000$ iterations and we considered $10000$ as burn-in which we left out. After discarding the burnin-in, for the next $50000$ iterations we estimate the targer probability as, \[
    \widehat{P} \;=\; \frac{1}{T - T_{\text{burn}}}
    \sum_{t=T_{\text{burn}}+1}^{T} \mathbf{1}\!\bigl[\text{diag}(X^{(t)}) = \mathbf{1}\bigr].
    \]
Now for $Q1$, at each iteration the candidate $Y$ is generated by drawing all 16 entries independently from $\text{Bernoulli}(0.5)$, so we can write
\[
q_1(Y \mid X) \;=\; \prod_{i,j=1}^{4} \left(\tfrac{1}{2}\right) \;=\; \frac{1}{2^{16}}.
\]
 We set a seed at `12345678` for reproducibility. As becuase $q_1(Y\mid X) = q_1(X\mid Y)$, the proposal is symmetric we can actually simplify the MH acceptance probability to 
\[
\alpha(X, Y) \;=\; \min\!\left(1,\; \frac{\pi(Y)}{\pi(X)}\right).
\]
In our R code we worked on the log scale, so we accept $Y$ if $\log U < \log\pi(Y) - \log\pi(X)$, such that $U \sim \text{Uniform}(0,1)$.

Now for $Q2$, we have at each iterations one of the 16 entries being selected uniformly at random (with probability $1/16$) and set it to 0 or 1 with equal probability $1/2$, leaving all other entries totally unchanged. Foe example, if entry $(i,j)$ is chosen,
\[
Y_{kl} =
\begin{cases}
\text{Bernoulli}(0.5), & (k,l) = (i,j),\\
X_{kl}, & (k,l) \neq (i,j).
\end{cases}
\]
We note that even this is a symmetric proposal as the probability of choosing entry $(i,j)$ and setting it to a particular value is the same in both directions, so the acceptance probability again can be reduced to
\[
\alpha(X, Y) \;=\; \min\!\left(1,\; \frac{\pi(Y)}{\pi(X)}\right).
\]

## 2) 

From the table below, it is evident that the second algorithm , namely $Q2$ has a much faster mixing Markov Chain. We see that for that, for 60000 iterations with 10000 burn-ins, the effective sample size for $Q1$ is 51.6 whereas for $Q2$ its 970.8 for $\lambda=0.5$ and for $\lambda = 1$ we wet the effective samples sizes for $Q1$ and $Q2$ as 3.4 and 795.1 respective;y. The acceptance rate for $Q1$ is 0.008 and 0.0002 for $\lambda = 0.5,1$ respectively. For $Q2$, on the other hand has a much much higher acceptance rate of $0.6$ and $0.5$ for $\lambda = 0.5,1$ respectively. This shows that $Q1$ almost never accepts, especially for $\lambda =1$ making the chain extremely difficult to move. Further proof can be obtained by looking at the MCMC standard errors which are lower for $Q2$ (0.012 and 0.009 for the two $\lambda$ values in the respective orders) than $Q1$ (0.029 and 0.022 for the two $\lambda$ values in the respective orders). 


```{r}
TAB
```
To make our point more convincing, we can look at the ACF plots below. The first two ACF plots are for $\lambda =0.5$, precisely, first one for $Q1$ and the second for $Q2$. The next two are for $\lambda =1$ in the same order. We see that $Q1$ for both lamda values have high autocorrelation with its past values whereas for $Q2$ it decreases with increasing lags, this shows that for $Q1$ the chains will get stuck and will decay much more than in $Q2$ making it way faster.  



```{r}
par(mfrow = c(2,2), mar = c(4,4,4,1))

acf(as.numeric(OUT$runs[["0.5"]]$Q1$indicator),
    main = "ACF: Q1 (lambda = 0.5)")

acf(as.numeric(OUT$runs[["0.5"]]$Q2$indicator),
    main = "ACF: Q2 (lambda = 0.5)")

acf(as.numeric(OUT$runs[["1"]]$Q1$indicator),
    main = "ACF: Q1 (lambda = 1)")

acf(as.numeric(OUT$runs[["1"]]$Q2$indicator),
    main = "ACF: Q2 (lambda = 1)")
```


## 3) 

Here I am sharing all the 12 plots (6 for $\lambda=0.5$ and 6 for $\lambda=1$) needed to answer this question. For $\lambda =1$, we see that under $Q1$, the trace plot of $I_t$ shows long flat line at 1, showing that the chain becomes stuck in high-probability configurations. This is consistent with the extremely low acceptance rate (0.0002) and very very small ESS (3.4). However for $q2$, we can see that the trace plot shows continuous switching between 0 and 1, which proves that it is the space. Similar conclusion can be drawn for both for $\lambda=0.5$ For the trace of D(X), we see for $\lambda = 1$, $Q1$ quickly collapses  with very small D(X) and remains there whereas $Q2$ again explors a wider range. Similar observations can be seen for the case of $\lambda =0.5$ but with far less extreme behavior. For the autocorrelation plots, as explained before, $Q14$ shows high autocorrelation with its lagged values than $Q2$ indicating higher changes of chain being stuck. To decide on the effective sample size, we looked at the effective sample size, the 	Batch-means MCMC standard errors which we recorded in the next subpart. All these indicates that $Q2$ is a much much faster mixing MC than $Q1$, thus we are suured that $Q1$ will converge way faster than $Q1$. Here comes a trade off, running too many iterations will be useless for $Q2$ as it has possible converged way earlier and will lead to unnecessary computation time. On the other hand, if we stop as soon as $Q2$ converged, there will be a good chance that $Q1$ haven't converged at all. So a trade of can be made by looking at the mean plots (estimate vs. iteration). There if we see that $Q1$ is about to or just (kind of) convered to almost fixed estimated values and $Q2$ converges already before a few iteration, that would, we believe, an excellent place to stop. By looking at the running mean plots below, it looks like at 50000 iterations after a burn-in of 10000 looks exactly what we wanted, $Q1$ is alomst stable at running mean of $0.87$ and $Q2$ have already stabilized at $0.9$.



```{r}
make_plots(OUT$runs[["0.5"]]$Q1, OUT$runs[["0.5"]]$Q2, main_tag="(lambda=0.5)")
make_plots(OUT$runs[["1"]]$Q1, OUT$runs[["1"]]$Q2, main_tag="(lambda=1)")
```
```{r}
running_mean <- function(x){
  x <- as.numeric(x)
  cumsum(x) / seq_along(x)
}

running_mean_plots <- function(res_Q1, res_Q2, lambda){

  ind1 <- as.numeric(res_Q1$indicator)
  ind2 <- as.numeric(res_Q2$indicator)

  rm1 <- running_mean(ind1)
  rm2 <- running_mean(ind2)

  par(mfrow=c(1,2), mar=c(4,4,4,1))

  plot(rm1, type="l",
       xlab="Iteration (post-burn)",
       ylab="Running mean",
       main=paste0("Running mean: Q1 (lambda=", lambda, ")"))
  abline(h=mean(ind1), lty=3)

  plot(rm2, type="l",
       xlab="Iteration (post-burn)",
       ylab="Running mean",
       main=paste0("Running mean: Q2 (lambda=", lambda, ")"))
  abline(h=mean(ind2), lty=3)
}

# Run for both lambdas
running_mean_plots(OUT$runs[["0.5"]]$Q1,
                   OUT$runs[["0.5"]]$Q2,
                   0.5)

running_mean_plots(OUT$runs[["1"]]$Q1,
                   OUT$runs[["1"]]$Q2,
                   1)
```


## 4)

The following Table reports all estimates, along with associated MCMC standard errors (using consistent mini batches). 

```{r}
TAB
```

From the table above me see that, $Q1$ has 0.4619 and 0.8483 as estimates and 0.0296 and 0.02270 as standard errors for $\lambda$ at 0.5 and 1 respectively. Similarly $Q3$ has 0.7707 and 0.9190 as estimates and 0.01265 and 0.00918 as standard errors for $\lambda$ at 0.5 and 1 respectively.






