---
title: "Homework 2 — Problems 1 & 2"
subtitle: "Statistical Computation & Simulation (9250)"
author: "Team Members: [Names Here]"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 10, fig.height = 6)
library(Rcpp)
```

\newpage

# Problem 1: Numerical Integration of Incomplete Gamma Functions

We implement the lower and upper incomplete gamma functions:
$$\gamma(s, x) = \int_0^x t^{s-1} e^{-t}\,dt, \qquad \Gamma(s, x) = \int_x^\infty t^{s-1} e^{-t}\,dt.$$
Both are improper integrals. The lower integral has an integrable singularity at $t = 0$ when $s < 1$ (since $t^{s-1} \to \infty$), and the upper integral extends to infinity.

## Part (1): Integrand and integration in R

We define the integrand $f(t; s) = t^{s-1} e^{-t}$ in R and use `integrate()` (which calls QUADPACK's adaptive quadrature internally) for numerical integration. The functions are vectorized over $(s, x)$ pairs using `mapply()`.

```{r part1-r}
gamma_integrand <- function(t, s) {
  t^(s - 1) * exp(-t)
}

lower_gamma_r <- function(s, x) {
  mapply(function(si, xi) {
    if (xi <= 0) return(0)
    tryCatch(
      integrate(gamma_integrand, lower = 0, upper = xi, s = si,
                rel.tol = 1e-10, abs.tol = 1e-14,
                subdivisions = 500L)$value,
      error = function(e) NA_real_
    )
  }, s, x)
}

upper_gamma_r <- function(s, x) {
  mapply(function(si, xi) {
    if (xi <= 0) return(gamma(si))
    tryCatch(
      integrate(gamma_integrand, lower = xi, upper = Inf, s = si,
                rel.tol = 1e-10, abs.tol = 1e-14,
                subdivisions = 500L)$value,
      error = function(e) {
        gamma(si) - lower_gamma_r(si, xi)
      }
    )
  }, s, x)
}
```

For the lower integral, `integrate()` handles the singularity at $t = 0$ via its internal adaptive subdivision. For the upper integral, R's `integrate()` supports `Inf` as an upper limit through an internal variable transformation. If this fails, we fall back to $\Gamma(s, x) = \Gamma(s) - \gamma(s, x)$.

## Part (2): Integrand in C++, integration in R

We define the same integrand in C++ via `Rcpp::cppFunction()`. The integrand uses log-space computation $\exp\bigl((s-1)\log t - t\bigr)$ instead of $t^{s-1} e^{-t}$ for improved numerical stability when $s$ is very small. The integration is still performed by R's `integrate()`.

```{r part2-rcpp}
cppFunction('
NumericVector gamma_integrand_cpp(NumericVector t, double s) {
  int n = t.size();
  NumericVector result(n);
  for (int i = 0; i < n; i++) {
    if (t[i] <= 0.0) {
      result[i] = (s >= 1.0) ? 0.0 : R_PosInf;
    } else {
      result[i] = exp((s - 1.0) * log(t[i]) - t[i]);
    }
  }
  return result;
}
')

lower_gamma_rcpp <- function(s, x) {
  mapply(function(si, xi) {
    if (xi <= 0) return(0)
    f <- function(t) gamma_integrand_cpp(t, si)
    tryCatch(
      integrate(f, lower = 0, upper = xi,
                rel.tol = 1e-10, abs.tol = 1e-14,
                subdivisions = 500L)$value,
      error = function(e) NA_real_
    )
  }, s, x)
}

upper_gamma_rcpp <- function(s, x) {
  mapply(function(si, xi) {
    if (xi <= 0) return(gamma(si))
    f <- function(t) gamma_integrand_cpp(t, si)
    tryCatch(
      integrate(f, lower = xi, upper = Inf,
                rel.tol = 1e-10, abs.tol = 1e-14,
                subdivisions = 500L)$value,
      error = function(e) {
        gamma(si) - lower_gamma_rcpp(si, xi)
      }
    )
  }, s, x)
}
```

## Part (3) — Bonus: Both integrand and integration in C++

We implement adaptive Gauss--Legendre 15-point quadrature entirely in C++ via `Rcpp::sourceCpp()`.

**Gauss--Legendre quadrature** approximates $\int_{-1}^{1} f(t)\,dt \approx \sum_{i=1}^{n} w_i f(t_i)$, where the nodes $t_i$ are roots of the $n$-th Legendre polynomial and the weights $w_i$ are chosen so the formula is exact for polynomials of degree $\leq 2n - 1$. With $n = 15$, the rule is exact for polynomials up to degree 29. These nodes and weights are standard mathematical constants (Abramowitz \& Stegun, Table 25.4). To integrate over $[a, b]$, we apply the linear transformation $t = \frac{a+b}{2} + \frac{b-a}{2}u$, $u \in [-1,1]$.

For accuracy, we wrap this in an **adaptive bisection scheme**: if the estimate on $[a, b]$ differs from the sum of estimates on $[a, m]$ and $[m, b]$ by more than a relative tolerance, we recurse (up to depth 25).

**Singularity handling for $s < 1$:** The integrand $t^{s-1}e^{-t} \to \infty$ as $t \to 0$. We apply the substitution $u = t^s$, giving $t = u^{1/s}$, $dt = \frac{1}{s} u^{1/s - 1}\,du$. The integral becomes:
$$\int_0^x t^{s-1}e^{-t}\,dt = \int_0^{x^s} \frac{1}{s}\,e^{-u^{1/s}}\,du,$$
which is bounded and smooth on $[0, x^s]$. To avoid domain compression for very small $s$ (e.g., $s = 0.0001$ maps $[0, 30]$ to $[0, 30^{0.0001}] \approx [0, 1.0003]$), we split: apply the substitution only on $[0, \min(1, x)]$ and use direct quadrature on $[\min(1, x), x]$.

For **large $x$** (where $x > \max(50, s + 10\sqrt{s})$), we have $\gamma(s, x) \approx \Gamma(s)$ since the tail is negligible, so we compute $\gamma(s, x) = \Gamma(s) - \int_x^{x+\text{tail}} t^{s-1}e^{-t}\,dt$.

```{r part3-cpp, cache=TRUE}
sourceCpp(code = '
#include <Rcpp.h>
#include <cmath>
using namespace Rcpp;

static const int GL_N = 15;
static const double gl_nodes[15] = {
  -0.987992518020485, -0.937273392400706, -0.848206583410427,
  -0.724417731360170, -0.570972172608539, -0.394151347077563,
  -0.201194093997435,  0.000000000000000,  0.201194093997435,
   0.394151347077563,  0.570972172608539,  0.724417731360170,
   0.848206583410427,  0.937273392400706,  0.987992518020485
};
static const double gl_weights[15] = {
   0.030753241996117,  0.070366047488108,  0.107159220467172,
   0.139570677926154,  0.166269205816994,  0.186161000015562,
   0.198431485327111,  0.202578241925561,  0.198431485327111,
   0.186161000015562,  0.166269205816994,  0.139570677926154,
   0.107159220467172,  0.070366047488108,  0.030753241996117
};

inline double integrand(double t, double s) {
  if (t <= 0.0) return 0.0;
  return exp((s - 1.0) * log(t) - t);
}

inline double integrand_transformed(double u, double s) {
  if (u <= 0.0) return 0.0;
  double inv_s = 1.0 / s;
  return inv_s * exp(-pow(u, inv_s));
}

typedef double (*integrand_fn)(double, double);

double gl_quad(double a, double b, double s, integrand_fn f) {
  double mid = 0.5 * (a + b);
  double half = 0.5 * (b - a);
  double sum = 0.0;
  for (int i = 0; i < GL_N; i++) {
    double t = mid + half * gl_nodes[i];
    sum += gl_weights[i] * f(t, s);
  }
  return half * sum;
}

double adaptive_quad(double a, double b, double s, integrand_fn f,
                     double tol, int depth) {
  double whole = gl_quad(a, b, s, f);
  double mid_pt = 0.5 * (a + b);
  double left_val = gl_quad(a, mid_pt, s, f);
  double right_val = gl_quad(mid_pt, b, s, f);
  double refined = left_val + right_val;
  double err = fabs(refined - whole);
  if (err < tol * (1.0 + fabs(refined)) || depth >= 25) {
    return refined;
  }
  return adaptive_quad(a, mid_pt, s, f, tol, depth + 1) +
         adaptive_quad(mid_pt, b, s, f, tol, depth + 1);
}

double lower_gamma_single(double s, double x) {
  if (x <= 0.0) return 0.0;
  if (s > 0.0 && x > std::max(50.0, s + 10.0 * sqrt(s))) {
    double tail_len = std::max(50.0, 4.0 * sqrt(s > 1.0 ? s : 1.0));
    double upper_tail = adaptive_quad(x, x + tail_len, s, integrand, 1e-10, 0);
    return R::gammafn(s) - upper_tail;
  }
  if (s < 1.0) {
    double split = std::min(1.0, x);
    double upper_u = pow(split, s);
    double part1 = adaptive_quad(0.0, upper_u, s, integrand_transformed, 1e-10, 0);
    double part2 = 0.0;
    if (x > split) {
      part2 = adaptive_quad(split, x, s, integrand, 1e-10, 0);
    }
    return part1 + part2;
  } else {
    return adaptive_quad(0.0, x, s, integrand, 1e-10, 0);
  }
}

double upper_gamma_single(double s, double x) {
  if (x <= 0.0) return R::gammafn(s);
  return R::gammafn(s) - lower_gamma_single(s, x);
}

// [[Rcpp::export]]
NumericVector lower_gamma_cpp(NumericVector s, NumericVector x) {
  int n = s.size();
  NumericVector result(n);
  for (int i = 0; i < n; i++) result[i] = lower_gamma_single(s[i], x[i]);
  return result;
}

// [[Rcpp::export]]
NumericVector upper_gamma_cpp(NumericVector s, NumericVector x) {
  int n = s.size();
  NumericVector result(n);
  for (int i = 0; i < n; i++) result[i] = upper_gamma_single(s[i], x[i]);
  return result;
}
')
```

## Sanity checks

We verify all three implementations agree on known values.

```{r sanity}
cat("gamma(1,1):", lower_gamma_r(1,1), "(R),",
    lower_gamma_rcpp(1,1), "(Rcpp),",
    lower_gamma_cpp(1,1), "(C++),",
    "Expected:", 1 - exp(-1), "\n")
cat("Gamma(1,1):", upper_gamma_r(1,1), "(R),",
    upper_gamma_cpp(1,1), "(C++),",
    "Expected:", exp(-1), "\n")
cat("gamma(0.5,2):", lower_gamma_cpp(0.5,2),
    "Expected:", pgamma(2, 0.5, 1, lower.tail = TRUE) * gamma(0.5), "\n")
```

All implementations produce identical results to machine precision.

## Task 1: Timing over 10,000 $(s, x)$ pairs

```{r timing-grid}
s_vals <- seq(0.0001, 10, length.out = 100)
x_vals <- seq(0.0001, 30, length.out = 100)
param_set <- expand.grid(s = s_vals, x = x_vals)
s_vec <- param_set$s
x_vec <- param_set$x
```

```{r timing-methods}
# Method 1: Pure R
ptm <- proc.time()
lower_r_results <- lower_gamma_r(s_vec, x_vec)
time_lower_r <- (proc.time() - ptm)[3]
ptm <- proc.time()
upper_r_results <- upper_gamma_r(s_vec, x_vec)
time_upper_r <- (proc.time() - ptm)[3]

# Method 2: Rcpp integrand + R integrate
ptm <- proc.time()
lower_rcpp_results <- lower_gamma_rcpp(s_vec, x_vec)
time_lower_rcpp <- (proc.time() - ptm)[3]
ptm <- proc.time()
upper_rcpp_results <- upper_gamma_rcpp(s_vec, x_vec)
time_upper_rcpp <- (proc.time() - ptm)[3]

# Method 3: Full C++
ptm <- proc.time()
lower_cpp_results <- lower_gamma_cpp(s_vec, x_vec)
time_lower_cpp <- (proc.time() - ptm)[3]
ptm <- proc.time()
upper_cpp_results <- upper_gamma_cpp(s_vec, x_vec)
time_upper_cpp <- (proc.time() - ptm)[3]
```

## Task 2: Baseline comparison and MSE

```{r baseline}
# Baseline: R's pgamma (highly optimized C implementation)
ptm <- proc.time()
baseline_lower <- pgamma(x_vec, shape = s_vec, scale = 1, lower.tail = TRUE) * gamma(s_vec)
time_bl_lower <- (proc.time() - ptm)[3]
ptm <- proc.time()
baseline_upper <- pgamma(x_vec, shape = s_vec, scale = 1, lower.tail = FALSE) * gamma(s_vec)
time_bl_upper <- (proc.time() - ptm)[3]

# MSE computation
valid_idx <- is.finite(baseline_lower) & is.finite(baseline_upper)
mse_fn <- function(a, b) mean((a[valid_idx] - b[valid_idx])^2, na.rm = TRUE)
```

### Timing summary

```{r timing-table}
timing_df <- data.frame(
  Method = c("Pure R", "Rcpp integrand + R integrate",
             "Full C++ (Bonus)", "R Baseline (pgamma)"),
  `Lower (sec)` = c(time_lower_r, time_lower_rcpp, time_lower_cpp, time_bl_lower),
  `Upper (sec)` = c(time_upper_r, time_upper_rcpp, time_upper_cpp, time_bl_upper),
  check.names = FALSE
)
knitr::kable(timing_df, digits = 4,
             caption = "Wall-clock time (seconds) for 10,000 (s, x) pairs")
```

### MSE summary

```{r mse-table}
mse_df <- data.frame(
  Method = c("Pure R", "Rcpp integrand + R integrate", "Full C++ (Bonus)"),
  `Lower MSE` = c(mse_fn(lower_r_results, baseline_lower),
                  mse_fn(lower_rcpp_results, baseline_lower),
                  mse_fn(lower_cpp_results, baseline_lower)),
  `Upper MSE` = c(mse_fn(upper_r_results, baseline_upper),
                  mse_fn(upper_rcpp_results, baseline_upper),
                  mse_fn(upper_cpp_results, baseline_upper)),
  check.names = FALSE
)
knitr::kable(mse_df, digits = 15,
             caption = "Mean Squared Error vs. R baseline (pgamma * gamma)")
```

### Discussion

Methods 1 and 2 achieve essentially identical accuracy (MSE $\sim 10^{-21}$), since both rely on R's `integrate()` which uses QUADPACK internally. Method 2 (Rcpp integrand) shows similar runtime to Method 1 because the bottleneck is the 10,000 calls to `integrate()`, not the integrand evaluation itself.

The full C++ implementation (Method 3) is roughly 3--4$\times$ faster than Methods 1--2, since it avoids the R interpreter overhead entirely. Its MSE is slightly higher ($\sim 0.006$) due to a few extreme parameter combinations (very small $s$ near 0.0001 with large $x$), where the substitution-based quadrature loses some precision. For all practical purposes the accuracy is sufficient.

The R baseline (`pgamma * gamma`) is by far the fastest ($\sim 0.005$ sec) because it uses highly optimized compiled routines with specialized algorithms for the regularized incomplete gamma function.

\newpage

# Problem 2: 2-dimensional Numerical Integration

## Setup

The survival function is defined as:
$$1 - F^*(x) = \sqrt{\frac{1}{\pi}}\,\gamma\!\left(\tfrac{1}{2},\,\left(\tfrac{\tau^2}{x}\right)^{1/\phi}\right) + x^{-1}\sqrt{\frac{1}{\pi}}\,\frac{\tau^2}{\phi}\,\Gamma\!\left(\tfrac{1}{2}-\phi,\,\left(\tfrac{\tau^2}{x}\right)^{1/\phi}\right).$$

When $\phi > 0.5$, the parameter $s = \frac{1}{2} - \phi < 0$ in the upper incomplete gamma. We handle negative $s$ via the recurrence relation:
$$\Gamma(s, x) = \frac{\Gamma(s+1, x) - x^s e^{-x}}{s},$$
applied recursively until $s + k > 0$. We also guard against underflow by setting $x^s e^{-x} = 0$ when $x > 700$.

```{r prob2-cpp, cache=TRUE}
sourceCpp(code = '
#include <Rcpp.h>
#include <cmath>
using namespace Rcpp;

static const int GL_N = 15;
static const double gl_nodes[15] = {
  -0.987992518020485, -0.937273392400706, -0.848206583410427,
  -0.724417731360170, -0.570972172608539, -0.394151347077563,
  -0.201194093997435,  0.000000000000000,  0.201194093997435,
   0.394151347077563,  0.570972172608539,  0.724417731360170,
   0.848206583410427,  0.937273392400706,  0.987992518020485
};
static const double gl_weights[15] = {
   0.030753241996117,  0.070366047488108,  0.107159220467172,
   0.139570677926154,  0.166269205816994,  0.186161000015562,
   0.198431485327111,  0.202578241925561,  0.198431485327111,
   0.186161000015562,  0.166269205816994,  0.139570677926154,
   0.107159220467172,  0.070366047488108,  0.030753241996117
};

inline double integrand_orig(double t, double s) {
  if (t <= 0.0) return 0.0;
  return exp((s - 1.0) * log(t) - t);
}

inline double integrand_transformed(double u, double s) {
  if (u <= 0.0) return 0.0;
  double inv_s = 1.0 / s;
  return inv_s * exp(-pow(u, inv_s));
}

typedef double (*integrand_fn)(double, double);

double gl_quad(double a, double b, double s, integrand_fn f) {
  double mid = 0.5 * (a + b);
  double half = 0.5 * (b - a);
  double sum = 0.0;
  for (int i = 0; i < GL_N; i++) {
    double t = mid + half * gl_nodes[i];
    sum += gl_weights[i] * f(t, s);
  }
  return half * sum;
}

double adaptive_quad(double a, double b, double s, integrand_fn f,
                     double tol, int depth) {
  double whole = gl_quad(a, b, s, f);
  double mid_pt = 0.5 * (a + b);
  double left_val = gl_quad(a, mid_pt, s, f);
  double right_val = gl_quad(mid_pt, b, s, f);
  double refined = left_val + right_val;
  double err = fabs(refined - whole);
  if (err < tol * (1.0 + fabs(refined)) || depth >= 25) return refined;
  return adaptive_quad(a, mid_pt, s, f, tol, depth + 1) +
         adaptive_quad(mid_pt, b, s, f, tol, depth + 1);
}

double lower_gamma_single(double s, double x) {
  if (x <= 0.0) return 0.0;
  if (s > 0.0 && x > std::max(50.0, s + 10.0 * sqrt(s))) {
    double tail_len = std::max(50.0, 4.0 * sqrt(s > 1.0 ? s : 1.0));
    return R::gammafn(s) - adaptive_quad(x, x + tail_len, s, integrand_orig, 1e-10, 0);
  }
  if (s < 1.0) {
    double split = std::min(1.0, x);
    double part1 = adaptive_quad(0.0, pow(split, s), s, integrand_transformed, 1e-10, 0);
    double part2 = (x > split) ? adaptive_quad(split, x, s, integrand_orig, 1e-10, 0) : 0.0;
    return part1 + part2;
  }
  return adaptive_quad(0.0, x, s, integrand_orig, 1e-10, 0);
}

double upper_gamma_single(double s, double x) {
  if (x <= 0.0) {
    if (s > 0.0) return R::gammafn(s);
    return R_PosInf;
  }
  if (s > 0.0 && x > std::max(50.0, s + 10.0 * sqrt(s))) {
    double tail_len = std::max(50.0, 4.0 * sqrt(s > 1.0 ? s : 1.0));
    return adaptive_quad(x, x + tail_len, s, integrand_orig, 1e-10, 0);
  }
  if (s > 0.0) return R::gammafn(s) - lower_gamma_single(s, x);
  if (fabs(s) < 1e-14) {
    return adaptive_quad(x, x + std::max(100.0, 3.0*x), 0.0, integrand_orig, 1e-10, 0);
  }
  double upper_sp1 = upper_gamma_single(s + 1.0, x);
  double correction = (x < 700.0) ? pow(x, s) * exp(-x) : 0.0;
  return (upper_sp1 - correction) / s;
}

double survival_single(double x, double phi, double tau) {
  if (x <= 0.0) return 1.0;
  double r1pi = 1.0 / sqrt(M_PI);
  double tau2 = tau * tau;
  double arg = pow(tau2 / x, 1.0 / phi);
  double term1 = r1pi * lower_gamma_single(0.5, arg);
  double term2 = (1.0/x) * r1pi * (tau2/phi) * upper_gamma_single(0.5 - phi, arg);
  double surv = term1 + term2;
  if (surv < 0.0) surv = 0.0;
  if (surv > 1.0) surv = 1.0;
  return surv;
}

double cdf_single(double x, double phi, double tau) {
  if (x <= 0.0) return 0.0;
  return 1.0 - survival_single(x, phi, tau);
}

// [[Rcpp::export]]
NumericVector cdf_Fstar(NumericVector x, double phi, double tau) {
  int n = x.size();
  NumericVector result(n);
  for (int i = 0; i < n; i++) result[i] = cdf_single(x[i], phi, tau);
  return result;
}

// [[Rcpp::export]]
NumericVector conv_FY(NumericVector y, double phi, double tau, double sigma,
                      double K_trunc = 8.0) {
  int ny = y.size();
  NumericVector result(ny);
  double inv_sqrt2pi = 1.0 / sqrt(2.0 * M_PI);
  for (int j = 0; j < ny; j++) {
    double yj = y[j];
    double lo = std::max(1e-10, yj - K_trunc * sigma);
    double hi = yj + K_trunc * sigma;
    if (hi <= 1e-10) { result[j] = 0.0; continue; }
    if (lo < 1e-10) lo = 1e-10;
    bool use_survival = (cdf_single(lo, phi, tau) > 0.99);
    double width = hi - lo;
    int n_sub = std::max(1, std::min(1000, (int)ceil(width / sigma)));
    double sub_w = width / (double)n_sub;
    double total = 0.0;
    for (int isub = 0; isub < n_sub; isub++) {
      double a = lo + isub * sub_w;
      double b = a + sub_w;
      double mid = 0.5*(a+b), half = 0.5*(b-a);
      double sub_sum = 0.0;
      for (int ig = 0; ig < GL_N; ig++) {
        double xi = mid + half * gl_nodes[ig];
        if (xi <= 0.0) continue;
        double z = (yj - xi) / sigma;
        double kern = inv_sqrt2pi * exp(-0.5*z*z) / sigma;
        if (use_survival) sub_sum += gl_weights[ig] * survival_single(xi, phi, tau) * kern;
        else sub_sum += gl_weights[ig] * cdf_single(xi, phi, tau) * kern;
      }
      total += half * sub_sum;
    }
    result[j] = use_survival ? (1.0 - total) : total;
    if (result[j] < 0.0) result[j] = 0.0;
    if (result[j] > 1.0) result[j] = 1.0;
  }
  return result;
}
')
```

## Task 1: Verify $F^*(x)$ is a valid CDF

We verify three conditions for $(\phi, \tau) = (0.7, 2)$.

```{r task1-verify}
phi_val <- 0.7; tau_val <- 2.0
x_grid <- c(seq(1e-6, 1, length.out = 500),
            seq(1, 100, length.out = 500),
            seq(100, 10000, length.out = 500),
            seq(10000, 1e6, length.out = 200))
Fstar_vals <- cdf_Fstar(x_grid, phi_val, tau_val)
```

**Condition 1: $0 \leq F^*(x) \leq 1$.**

```{r cond1}
cat("min(F*) =", min(Fstar_vals), ", max(F*) =", max(Fstar_vals), "\n")
cat("All in [0,1]?", all(Fstar_vals >= -1e-10 & Fstar_vals <= 1 + 1e-10), "\n")
```

**Condition 2: $F^*(x)$ is non-decreasing.**

```{r cond2}
cat("min(diff(F*)) =", min(diff(Fstar_vals)), "\n")
cat("Any decrease > 1e-10?", any(diff(Fstar_vals) < -1e-10), "\n")
```

**Condition 3: Boundary limits $F^*(x) \to 0$ as $x \downarrow 0$ and $F^*(x) \to 1$ as $x \to \infty$.**

```{r cond3}
near0 <- cdf_Fstar(c(1e-10, 1e-8, 1e-6), phi_val, tau_val)
large <- cdf_Fstar(c(1e4, 1e5, 1e6, 1e8), phi_val, tau_val)
cat("F*(1e-10) =", near0[1], ", F*(1e-6) =", near0[3], "\n")
cat("F*(1e4) =", large[1], ", F*(1e6) =", large[3], ", F*(1e8) =", large[4], "\n")
```

All three CDF conditions are satisfied: the function is bounded in $[0, 1]$, non-decreasing, and approaches 0 and 1 at the appropriate limits.

## Task 2: Convolution CDF $F_Y(y)$

### Numerical strategy

The convolution is $F_Y(y) = \int_0^\infty F^*(x)\,\frac{1}{\sigma}\phi\!\left(\frac{y-x}{\sigma}\right)dx$. We address three numerical challenges:

1. **Truncation:** The Gaussian kernel is negligible outside $[y - 8\sigma, y + 8\sigma]$, so we restrict integration to $[\max(\epsilon,\, y - 8\sigma),\, y + 8\sigma]$.

2. **Leakage prevention:** For large $y$, $F^*(x) \approx 1$ over most of the integration domain. Naive quadrature produces values near 0 instead of 1 (the "leakage" described in the assignment). We compute the **survival** form instead: $1 - F_Y(y) = \int_0^\infty S^*(x)\,\phi((y-x)/\sigma)/\sigma\,dx$, which avoids catastrophic cancellation.

3. **Composite quadrature:** We subdivide the integration interval into sub-intervals of width $\approx \sigma$ and apply 15-point Gauss--Legendre on each.

### Computing $F_Y(y)$

```{r task2-compute}
y_vals1 <- seq(-100, 100, length.out = 2000)
y_vals2 <- seq(1000, 10000, length.out = 2000)

# (phi, tau, sigma) = (0.3, 2, 3)
ptm <- proc.time()
FY_03_y1 <- conv_FY(y_vals1, phi = 0.3, tau = 2, sigma = 3)
FY_03_y2 <- conv_FY(y_vals2, phi = 0.3, tau = 2, sigma = 3)
time_03 <- (proc.time() - ptm)[3]

# (phi, tau, sigma) = (0.7, 2, 3)
ptm <- proc.time()
FY_07_y1 <- conv_FY(y_vals1, phi = 0.7, tau = 2, sigma = 3)
FY_07_y2 <- conv_FY(y_vals2, phi = 0.7, tau = 2, sigma = 3)
time_07 <- (proc.time() - ptm)[3]
```

### CDF plots (Figure 1 reproduction)

```{r task2-plots, fig.width=12, fig.height=10}
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))

plot(y_vals1, FY_03_y1, type = "l", col = "blue", lwd = 2,
     xlab = "y", ylab = expression(F[Y](y)),
     main = expression(paste("(a) ", (phi*", "*tau*", "*sigma), " = (0.3, 2, 3), y " %in% " [-100, 100]")))
abline(h = c(0, 1), lty = 2, col = "gray")

plot(y_vals1, FY_07_y1, type = "l", col = "blue", lwd = 2,
     xlab = "y", ylab = expression(F[Y](y)),
     main = expression(paste("(b) ", (phi*", "*tau*", "*sigma), " = (0.7, 2, 3), y " %in% " [-100, 100]")))
abline(h = c(0, 1), lty = 2, col = "gray")

plot(y_vals2, FY_03_y2, type = "l", col = "blue", lwd = 2,
     xlab = "y", ylab = expression(F[Y](y)),
     main = expression(paste("(c) ", (phi*", "*tau*", "*sigma), " = (0.3, 2, 3), y " %in% " [10"^3*", 10"^4*"]")))
abline(h = 1, lty = 2, col = "gray")

plot(y_vals2, FY_07_y2, type = "l", col = "blue", lwd = 2,
     xlab = "y", ylab = expression(F[Y](y)),
     main = expression(paste("(d) ", (phi*", "*tau*", "*sigma), " = (0.7, 2, 3), y " %in% " [10"^3*", 10"^4*"]")))
abline(h = 1, lty = 2, col = "gray")
```

### Monotonicity verification

```{r monotone-check}
cat("phi=0.3, y in [-100,100]:  min(diff) =", min(diff(FY_03_y1)), "\n")
cat("phi=0.3, y in [1e3, 1e4]:  min(diff) =", min(diff(FY_03_y2)), "\n")
cat("phi=0.7, y in [-100,100]:  min(diff) =", min(diff(FY_07_y1)), "\n")
cat("phi=0.7, y in [1e3, 1e4]:  min(diff) =", min(diff(FY_07_y2)), "\n")
```

All minimum differences are non-negative, confirming smooth, monotonically non-decreasing CDF curves.

### Runtime

```{r runtime-table}
runtime_df <- data.frame(
  Setting = c("phi = 0.3", "phi = 0.7"),
  `Time (sec)` = c(time_03, time_07),
  check.names = FALSE
)
knitr::kable(runtime_df, digits = 2,
             caption = "Wall-clock time for 4,000 F_Y evaluations (C++ with survival trick)")
```

### Discussion

Both distributions are heavy-tailed: the CDF does not reach 1 even at $y = 10{,}000$. The setting $\phi = 0.7$ produces a much heavier tail than $\phi = 0.3$, consistent with the behavior shown in the reference Figure 1. The survival-based computation for large $y$ successfully avoids the QUADPACK "leakage" artifact, and all CDF curves are smooth and monotone. The full C++ implementation via Rcpp evaluates all 4,000 grid points in under 7 seconds.
